---
title: "Analysing the outcomes of Legendplex data"
author: "Sidney van der Zande"
date: "`25-07-2024`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, collapse=T)
```


## Notes
This script is meant to analyse the output of the Legendplex data for all tumours, generated in the Legendplex.Rmd script. From the windsor-normalised data, it generates heatmaps on a per tumour, condition, segment level and so on.

As an aditional note, while the segment types are here named "TLS", "intermediate" and "tumour"/"tumour bed", these names were later changed in the graphs and report to "B cell high", "B cell intermediate" and "B cell low", respectively.

## 1.Generation of data
```{r Load libraries}
#Load libraries
source("~/Documents/R scripts/Functions.R") #this is a script which contains functions specifically written for the TLS and SLICE projects
library(readxl)
library(openxlsx)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(forstringr)
```

```{r Read in data from Legendplex.Rmd file}
#read in normalised data
lu023_TLS014 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU023_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu023_TLS014) <- lu023_TLS014[,1]
lu023_TLS014[,1] <- NULL
rownames(lu023_TLS014) <- gsub("Unstim", "unstim", rownames(lu023_TLS014))

lu088 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU088_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu088) <- lu088[,1]
lu088[,1] <- NULL
rownames(lu088) <- gsub("Unstim", "unstim", rownames(lu088))

lu091 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU091_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu091) <- lu091[,1]
lu091[,1] <- NULL
rownames(lu091) <- gsub("Unstim", "unstim", rownames(lu091))

lu035_TLS013 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/LU035_TLS013_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu035_TLS013) <- lu035_TLS013[,1]
lu035_TLS013[,1] <- NULL

lu035_TLS015 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/LU035_TLS015_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu035_TLS015) <- lu035_TLS015[,1]
lu035_TLS015[,1] <- NULL

lu066_TLS015 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU066_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu066_TLS015) <- lu066_TLS015[,1]
lu066_TLS015[,1] <- NULL

lu052_TLS010 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU052_TLS010_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu052_TLS010) <- lu052_TLS010[,1]
lu052_TLS010[,1] <- NULL
rownames(lu052_TLS010) <- gsub("hrIFNy", "hrIFNy_high", rownames(lu052_TLS010))

lu052_TLS014 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU052_TLS014_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu052_TLS014) <- lu052_TLS014[,1]
lu052_TLS014[,1] <- NULL

lu023_TLS013 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU023_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu023_TLS013) <- lu023_TLS013[,1]
lu023_TLS013[,1] <- NULL

lu086_2 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU086-2_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu086_2) <- lu086_2[,1]
lu086_2[,1] <- NULL

lu083_TLS015 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU083_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu083_TLS015) <- lu083_TLS015[,1]
lu083_TLS015[,1] <- NULL

lu086_1 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU086-1_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu086_1) <- lu086_1[,1]
lu086_1[,1] <- NULL

lu066_TLS012 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU066_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu066_TLS012) <- lu066_TLS012[,1]
lu066_TLS012[,1] <- NULL

lu083_TLS012 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU083_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu083_TLS012) <- lu083_TLS012[,1]
lu083_TLS012[,1] <- NULL

lu080 <- as.data.frame(read_excel("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU080_normalised_windsor.xlsx", sheet = 3)) #corrected values
rownames(lu080) <- lu080[,1]
lu080[,1] <- NULL
```

```{r Read in whether each fragment is TLS, intermediate or tumour bed; FlowJo_TLS.Rmd file}
#read in TLS information
TLS_status <- read.csv("~/Documents/Analyses/TLS/fragment_table.csv") #generated in the file FlowJo_TLS.Rmd
TLS_status$X <- NULL
TLS_status$heatmap <- paste0(TLS_status$tumour, " ", TLS_status$condition, " fragment " ,TLS_status$fragment) #to couple TLS information to fragments in Legendplex
TLS_status$heatmap <- gsub("aIFNyR1\\+aPD1", "aPD1+aIFNyR1", TLS_status$heatmap) #rename incorrectly names conditions

#For Lu035 and LU052, the fragment is named with the experiment number in the Legendplex data. Adjust this in the FlowJo data as well to match
TLS_status$heatmap[TLS_status$experiment == "TLS013" & TLS_status$tumour == "LU035"] <- paste0("TLS013 ", TLS_status$heatmap[TLS_status$experiment == "TLS013" & TLS_status$tumour == "LU035"])
TLS_status$heatmap[TLS_status$experiment == "TLS015" & TLS_status$tumour == "LU035"] <- paste0("TLS015 ", TLS_status$heatmap[TLS_status$experiment == "TLS015" & TLS_status$tumour == "LU035"])
TLS_status$heatmap[TLS_status$experiment == "TLS010" & TLS_status$tumour == "LU052"] <- paste0("TLS010 ", TLS_status$heatmap[TLS_status$experiment == "TLS010" & TLS_status$tumour == "LU052"])
TLS_status$heatmap[TLS_status$experiment == "TLS014" & TLS_status$tumour == "LU052"] <- paste0("TLS014 ", TLS_status$heatmap[TLS_status$experiment == "TLS014" & TLS_status$tumour == "LU052"])

#denote if LU086 is replicate 1 or 2
TLS_status$heatmap[TLS_status$experiment == "TLS011" & grepl("LU086_1",TLS_status$Name)] <- gsub("LU086", "LU086-1", TLS_status$heatmap[TLS_status$experiment == "TLS011" & grepl("LU086_1",TLS_status$Name)])
TLS_status$heatmap[TLS_status$experiment == "TLS011" & grepl("LU086_2",TLS_status$Name)] <- gsub("LU086", "LU086-2", TLS_status$heatmap[TLS_status$experiment == "TLS011" & grepl("LU086_2",TLS_status$Name)])

#remove tumours that were not assessed (at this point only TLS009 LU086)
TLS_status <- TLS_status[TLS_status$experiment != "TLS009",]

#remove the RN440-15 and RN440-16 conditions
TLS_status <- TLS_status[!(TLS_status$condition %in% c("RN440-15", "RN440-16")),]
```

```{r Separate TLS, intermediate and tumour fragments per tumour}
#Get TLS segments out per tumour
TLS_023_013 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS013","heatmap"]
int_023_013 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS013","heatmap"]
tumour_023_013 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS013","heatmap"]

TLS_023_014 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS014","heatmap"]
int_023_014 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS014","heatmap"]
tumour_023_014 <- TLS_status[TLS_status$tumour == "LU023"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS014","heatmap"]

TLS_035_013 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS013","heatmap"]
int_035_013 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS013","heatmap"]
tumour_035_013 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS013","heatmap"]

TLS_035_015 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS015","heatmap"]
int_035_015 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS015","heatmap"]
tumour_035_015 <- TLS_status[TLS_status$tumour == "LU035"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS015","heatmap"]

TLS_052_010 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS010","heatmap"]
int_052_010 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS010","heatmap"]
tumour_052_010 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS010","heatmap"]

TLS_052_014 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS014","heatmap"]
int_052_014 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS014","heatmap"]
tumour_052_014 <- TLS_status[TLS_status$tumour == "LU052"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS014","heatmap"]

TLS_066_015 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS015","heatmap"]
int_066_015 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS015","heatmap"]
tumour_066_015 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS015","heatmap"]

TLS_066_012 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS012","heatmap"]
int_066_012 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS012","heatmap"]
tumour_066_012 <- TLS_status[TLS_status$tumour == "LU066"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS012","heatmap"]

TLS_080 <- TLS_status[TLS_status$tumour == "LU080" & TLS_status$segment == "TLS" ,"heatmap"]
int_080 <- TLS_status[TLS_status$tumour == "LU080"& TLS_status$segment == "intermediate","heatmap"]
tumour_080 <- TLS_status[TLS_status$tumour == "LU080"& TLS_status$segment == "tumour_bed","heatmap"]

TLS_083_012 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS012","heatmap"]
int_083_012 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS012","heatmap"]
tumour_083_012 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS012","heatmap"]

TLS_083_015 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "TLS" & TLS_status$experiment == "TLS015","heatmap"]
int_083_015 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS015","heatmap"]
tumour_083_015 <- TLS_status[TLS_status$tumour == "LU083"& TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS015","heatmap"]

TLS_086_1 <- TLS_status[grepl("LU086_1", TLS_status$Name) & TLS_status$segment == "TLS" & TLS_status$experiment == "TLS011","heatmap"]
int_086_1 <- TLS_status[grepl("LU086_1", TLS_status$Name) & TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS011","heatmap"]
tumour_086_1 <- TLS_status[grepl("LU086_1", TLS_status$Name) & TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS011","heatmap"]

TLS_086_2 <- TLS_status[grepl("LU086_2", TLS_status$Name) & TLS_status$segment == "TLS" & TLS_status$experiment == "TLS011","heatmap"]
int_086_2 <- TLS_status[grepl("LU086_2", TLS_status$Name) & TLS_status$segment == "intermediate" & TLS_status$experiment == "TLS011","heatmap"]
tumour_086_2 <- TLS_status[grepl("LU086_2", TLS_status$Name) & TLS_status$segment == "tumour_bed" & TLS_status$experiment == "TLS011","heatmap"]

TLS_088 <- TLS_status[TLS_status$tumour == "LU088" & TLS_status$segment == "TLS" ,"heatmap"]
int_088 <- TLS_status[TLS_status$tumour == "LU088"& TLS_status$segment == "intermediate","heatmap"]
tumour_088 <- TLS_status[TLS_status$tumour == "LU088"& TLS_status$segment == "tumour_bed","heatmap"]

TLS_091 <- TLS_status[TLS_status$tumour == "LU091" & TLS_status$segment == "TLS" ,"heatmap"]
int_091 <- TLS_status[TLS_status$tumour == "LU091"& TLS_status$segment == "intermediate","heatmap"]
tumour_091 <- TLS_status[TLS_status$tumour == "LU091"& TLS_status$segment == "tumour_bed","heatmap"]
```

## 2. Quality control

```{r Quality control}
#Run QCs (from source file Functions.R)

#make a list of all file paths containing the location to the data per tumour
file_list <- c("~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU023.xlsx", "~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU088.xlsx", 
"~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LU091.xlsx", "~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/LU035_TLS013.xlsx", 
"~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/LU035_TLS015.xlsx", "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU066.xlsx", 
"~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU052_TLS010.xlsx", "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LU052_TLS014.xlsx",
"~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU023_TLS013.xlsx","~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU083_TLS015.xlsx",
"~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LU086-2_TLS011.xlsx","~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU066.xlsx",
"~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU080.xlsx","~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU083.xlsx",
"~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU086-1.xlsx") 

#do the same for the files containing the MFI for each Legendplex (duplicated because each Legendplex contains multiple tumours)
MFI_file_list <- c("~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/MFI LP063.xlsx","~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/MFI LP063.xlsx",
                   "~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/MFI LP063.xlsx","~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/MFI LP064.xlsx",
                   "~/Documents/Experiments/TLS/LP064-Timm-LU035(2)/MFI LP064.xlsx","~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/MFI LP065.xlsx",
                   "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/MFI LP065.xlsx","~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/MFI LP065.xlsx",
                   "~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LP066_MFI.xlsx","~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LP066_MFI.xlsx",
                   "~/Documents/Experiments/TLS/LP066-LU023-LU083-LU086/LP066_MFI.xlsx","~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LP060_MFI.xlsx",
                   "~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LP060_MFI.xlsx","~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LP060_MFI.xlsx",
                   "~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LP060_MFI.xlsx")

#For each entry in the file, add information on the tumour, experiment and Legendplex number
tumours <- c("LU023", "LU088", "LU091", "LU035", "LU035", "LU066", "LU052", "LU052", "LU023", "LU083", "LU086-2", "LU066", "LU080", "LU083", "LU086-1")
experiments <- c("TLS014", "TLS016", "TLS016", "TLS013", "TLS015", "TLS015", "TLS010", "TLS014" ,"TLS013","TLS015","TLS011","TLS012","TLS011","TLS012","TLS011")
legendplex <- c("LP063","LP063","LP063","LP064","LP064","LP065","LP065","LP065","LP066","LP066","LP066","LP060","LP060","LP060","LP060")

conc_plots <- list()
MFI_plots <- list()
stat_list <- list()

for (i in seq_along(file_list)){ #for each tumour
  pdf(paste0("~/Documents/Analyses/TLS/QC plots/QC_", tumours[[i]], "_",experiments[[i]],".pdf"), width = 15) #make a PDF file conaining the QC for that tumour
  print((paste0(tumours[[i]], " ", experiments[[i]], " report"))) #name the tumour that is being investigated
  
  #do QC based on comparing concentrations to unstim
  cat("Based on concentration:","\n")
  conc_plots[[length(conc_plots)+1]] <- legendplex_qc_concentration(file_list[[i]]) #from the file Functions.R
  names(conc_plots)[[length(conc_plots)]] <- paste0(tumours[[i]], "_", experiments[[i]]) #save to list
  print(conc_plots[[i]])
  
  #do QC based on comparing MFI to unstim
  cat("Based on MFI:","\n")
  MFI_plots[[length(MFI_plots)+1]] <- legendplex_qc_MFI(MFI_file = MFI_file_list[[i]], experiment = experiments[[i]], tumour_name = tumours[[i]]) #from the file Functions.R
  names(MFI_plots)[[length(MFI_plots)]] <- paste0(tumours[[i]], "_", experiments[[i]]) #save to list
  print(MFI_plots[[i]])
  
  dev.off()
  
  #save the fold-changes to a list
  stat_list[[length(stat_list)+1]] <- legendplex_fragment_control(file_list[[i]], experiment = experiments[[i]])
  names(stat_list)[[length(stat_list)]] <- paste0(tumours[[i]], "_", experiments[[i]])
}
```

## 3. Data without outlier removal

First, we generate a heatmap of mediator expression per segemnt type (TLS-int-tumour), averaged across all fragments, tumours and conditions.
```{r Segment-based heatmap}
#Get separate Legendplex scores per segment
TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,] #select legendplex entries of LU023 that match the fragments identified as TLS in the FlowJo data
TLS_023_013_data <- colMeans(TLS_023_013_data) #average score per cytokine across all conditions
int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,] #repeat for intermediate fragments
int_023_013_data <- colMeans(int_023_013_data)
tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,] #repeat for tumour bed fragments
tumour_023_013_data <- colMeans(tumour_023_013_data)

#repeat for LU023 in experiment TLS014
TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
TLS_023_014_data <- colMeans(TLS_023_014_data) #average score per cytokine across all conditions
int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
int_023_014_data <- colMeans(int_023_014_data)
tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]
tumour_023_014_data <- colMeans(tumour_023_014_data)

TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
TLS_035_013_data <- colMeans(TLS_035_013_data) #average score per cytokine across all conditions
int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
int_035_013_data <- colMeans(int_035_013_data)
tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]
tumour_035_013_data <- colMeans(tumour_035_013_data)

TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
TLS_035_015_data <- colMeans(TLS_035_015_data) #average score per cytokine across all conditions
int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
int_035_015_data <- colMeans(int_035_015_data)
tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]
tumour_035_015_data <- colMeans(tumour_035_015_data)

TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
TLS_052_010_data <- colMeans(TLS_052_010_data) #average score per cytokine across all conditions
int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
int_052_010_data <- colMeans(int_052_010_data)
tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
tumour_052_010_data <- colMeans(tumour_052_010_data)

TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
TLS_052_014_data <- colMeans(TLS_052_014_data) #average score per cytokine across all conditions
int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
int_052_014_data <- colMeans(int_052_014_data)
tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]
tumour_052_014_data <- colMeans(tumour_052_014_data)

#This tumour has no intermediate fragments. Therefore, this condition is skipped.
TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
TLS_066_012_data <- colMeans(TLS_066_012_data) #average score per cytokine across all conditions
tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]
tumour_066_012_data <- colMeans(tumour_066_012_data)

TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
TLS_066_015_data <- colMeans(TLS_066_015_data) #average score per cytokine across all conditions
int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
int_066_015_data <- colMeans(int_066_015_data)
tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]
tumour_066_015_data <- colMeans(tumour_066_015_data)

#LU080 has no TLS so we ignore this condition
int_080_data <- lu080[rownames(lu080) %in% int_080,]
int_080_data <- colMeans(int_080_data)
tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]
tumour_080_data <- colMeans(tumour_080_data)

TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
TLS_083_012_data <- colMeans(TLS_083_012_data) #average score per cytokine across all conditions
int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
int_083_012_data <- colMeans(int_083_012_data)
tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]
tumour_083_012_data <- colMeans(tumour_083_012_data)

TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
TLS_083_015_data <- colMeans(TLS_083_015_data) #average score per cytokine across all conditions
int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
int_083_015_data <- colMeans(int_083_015_data)
tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]
tumour_083_015_data <- colMeans(tumour_083_015_data)

#the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
int_086_1_data <- colMeans(int_086_1_data)
tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]
tumour_086_1_data <- colMeans(tumour_086_1_data)

#the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
int_086_2_data <- colMeans(int_086_2_data)
tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]
tumour_086_2_data <- colMeans(tumour_086_2_data)

TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
TLS_088_data <- colMeans(TLS_088_data) #average score per cytokine across all conditions
int_088_data <- lu088[rownames(lu088) %in% int_088,]
int_088_data <- colMeans(int_088_data)
tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]
tumour_088_data <- colMeans(tumour_088_data)

TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
TLS_091_data <- colMeans(TLS_091_data) #average score per cytokine across all conditions
int_091_data <- lu091[rownames(lu091) %in% int_091,]
int_091_data <- colMeans(int_091_data)
tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]
tumour_091_data <- colMeans(tumour_091_data)

#Average across all tumours
TLS_scores <- rbind(TLS_023_013_data,TLS_023_014_data, TLS_035_013_data, TLS_035_015_data, TLS_052_010_data, TLS_052_014_data, TLS_066_012_data,TLS_066_015_data,
                    TLS_083_012_data, TLS_083_015_data, TLS_088_data, TLS_091_data) #missing LU080, LU086-1, LU086-2
TLS_scores <- colMeans(TLS_scores)

int_scores <- rbind(int_023_013_data,int_023_014_data, int_035_013_data, int_035_015_data, int_052_010_data, int_052_014_data, int_066_015_data,
                    int_080_data, int_083_012_data, int_083_015_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data) #missing LU066 (TLS012)
int_scores <- colMeans(int_scores)

tumour_scores <- rbind(tumour_023_013_data,tumour_023_014_data, tumour_035_013_data, tumour_035_015_data, tumour_052_010_data, tumour_052_014_data, tumour_066_012_data,tumour_066_015_data,
                       tumour_080_data, tumour_083_012_data, tumour_083_015_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data)
tumour_scores <- colMeans(tumour_scores)

heatmap_scores <- rbind(TLS_scores, int_scores, tumour_scores)
rownames(heatmap_scores) <- c("TLS", "intermediate", "tumour bed")

#Generate a heatmap of z-scores per cytokine
library(pheatmap)
  #z-score is (value - average)/standard deviation. 
  colSD <- apply(heatmap_scores, 2, sd) #calculate the standard deviation per cytokine
  colMean <- colMeans(heatmap_scores) #calculate the average per cytokine
  z_scores <- sweep(heatmap_scores, 2, colMean, "-") #subtract means per column
  z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
  z_scores <- t(z_scores) #transpoe for the heatmap
  
  rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"
  rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y"
  
  #heatmap
  heatmap <- suppressWarnings(pheatmap(z_scores))  #make a heatmap
  heatmap
  
  
  save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_normalised_windsor.pdf") #save to PDF
```

Now, we generate a heatmap which is split per segment type as well as condition (TLS-unstim, TLS-aPD1 etc.)
```{r Segment/condition based heatmap}
  #split per condition
  #Again, subset the TLS, intermediate and tumour bed fragments per tumour, but do not take the averages per tumour
  TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,]
  int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,]
  tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,]

  TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
  int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
  tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]

  TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
  int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
  tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]

  TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
  int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
  tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]

  TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
  int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
  tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
  #LU052_TLS010_tumour_bed has extremely high CXCL9/CXCL10, what does the heatmap look like if I take the IFN_high fragments out?
  TLS_052_010_data <- TLS_052_010_data[!grepl("hrIFNy_high", rownames(TLS_052_010_data)),]
  int_052_010_data <- int_052_010_data[!grepl("hrIFNy_high", rownames(int_052_010_data)),]
  tumour_052_010_data <- tumour_052_010_data[!grepl("hrIFNy_high", rownames(tumour_052_010_data)),]
  
  TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
  int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
  tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]

  #This tumour has no intermediate fragments. Therefore, this condition is skipped.
  TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
  tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]

  TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
  int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
  tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]

  #LU080 has no TLS so we ignore this condition
  int_080_data <- lu080[rownames(lu080) %in% int_080,]
  tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]

  TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
  int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
  tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]

  TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
  int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
  tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]

  #the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
  int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
  tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]

  #the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
  int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
  tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]

  TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
  int_088_data <- lu088[rownames(lu088) %in% int_088,]
  tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]

  TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
  int_091_data <- lu091[rownames(lu091) %in% int_091,]
  tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]

  #First for TLS
  TLS_list <- list(TLS_023_013_data,TLS_023_014_data, TLS_035_013_data, TLS_035_015_data, TLS_052_010_data, TLS_052_014_data, TLS_066_012_data,TLS_066_015_data,
                   TLS_083_012_data, TLS_083_015_data, TLS_088_data, TLS_091_data) #missing LU080, LU086-1, LU086-2
  names(TLS_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012","LU066_TLS015",
                       "LU083_TLS012", "LU083_TLS015", "LU088", "LU091")  
  
  #make data frames for each condition
  unstim_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25)) #25 columns because there are 25 mediators in the panel
  aPD1_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Calculate condition averages per tumour
  for(i in seq_along(TLS_list)){ #for each tumour (TLS fragments only)
    data <- TLS_list[[i]] #grab the data
    
    #split fragments across conditions
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),] #grab fragments with 'unstim' in their name
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    unstim <- colMeans(unstim) #calculate the mean cytokine/chemokine concentration per condition for this tumour
    aPD1 <- colMeans(aPD1)
    aPD1_aIFNR <- colMeans(aPD1_aIFNR)
    hrIFNy <- colMeans(hrIFNy)
    
    if(!any(grepl("NaN", unstim))){ #if there is an unstim condition for this tumour,add this entry to the list
    unstim_average_TLS <- rbind(unstim_average_TLS, unstim)
    rownames(unstim_average_TLS)[[nrow(unstim_average_TLS)]] <- names(TLS_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1))){ #if there is an aPD1 condition for this tumour,add this entry to the list
      aPD1_average_TLS <- rbind(aPD1_average_TLS, aPD1)
      rownames(aPD1_average_TLS)[[nrow(aPD1_average_TLS)]] <- names(TLS_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1_aIFNR))){ #if there is an aPD1+aIFNyR1 condition for this tumour,add this entry to the list
      aPD1_aIFNR_average_TLS <- rbind(aPD1_aIFNR_average_TLS, aPD1_aIFNR)
      rownames(aPD1_aIFNR_average_TLS)[[nrow(aPD1_aIFNR_average_TLS)]] <- names(TLS_list)[[i]]
    }
    
    if(!any(grepl("NaN", hrIFNy))){ #if there is an hrIFNy condition for this tumour,add this entry to the list
      hrIFNy_average_TLS <- rbind(hrIFNy_average_TLS, hrIFNy)
      rownames(hrIFNy_average_TLS)[[nrow(hrIFNy_average_TLS)]] <- names(TLS_list)[[i]]
    }

  }
  
  #Average across all tumours, but save a backup for other heatmaps
  colnames(unstim_average_TLS) <- colnames(TLS_023_014_data)
  unstim_average_TLS_tumour <- unstim_average_TLS
  unstim_average_TLS <- colMeans(unstim_average_TLS)
  
  colnames(aPD1_average_TLS) <- colnames(TLS_023_014_data)
  aPD1_average_TLS_tumour <- aPD1_average_TLS
  aPD1_average_TLS <- colMeans(aPD1_average_TLS)
  
  colnames(aPD1_aIFNR_average_TLS) <- colnames(TLS_023_014_data)
  aPD1_aIFNR_average_TLS_tumour <- aPD1_aIFNR_average_TLS
  aPD1_aIFNR_average_TLS <- colMeans(aPD1_aIFNR_average_TLS)
  
  colnames(hrIFNy_average_TLS) <- colnames(TLS_023_014_data)
  hrIFNy_average_TLS_tumour <- hrIFNy_average_TLS
  hrIFNy_average_TLS <- colMeans(hrIFNy_average_TLS)
  
  #repeat for intermediate and tumour sections
  int_list <- list(int_023_013_data,int_023_014_data, int_035_013_data, int_035_015_data, int_052_010_data, int_052_014_data, int_066_015_data,
                   int_080_data, int_083_012_data, int_083_015_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data) #missing LU066 (TLS012)
  names(int_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014", "LU066_TLS015",
                       "LU080","LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
  unstim_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Calculate condition averages per tumour
  for(i in seq_along(int_list)){
    data <- int_list[[i]]
    
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    unstim <- colMeans(unstim)
    aPD1 <- colMeans(aPD1)
    aPD1_aIFNR <- colMeans(aPD1_aIFNR)
    hrIFNy <- colMeans(hrIFNy)
    
    if(!any(grepl("NaN", unstim))){
    unstim_average_int <- rbind(unstim_average_int, unstim)
    rownames(unstim_average_int)[[nrow(unstim_average_int)]] <- names(int_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1))){
      aPD1_average_int <- rbind(aPD1_average_int, aPD1)
      rownames(aPD1_average_int)[[nrow(aPD1_average_int)]] <- names(int_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1_aIFNR))){
      aPD1_aIFNR_average_int <- rbind(aPD1_aIFNR_average_int, aPD1_aIFNR)
      rownames(aPD1_aIFNR_average_int)[[nrow(aPD1_aIFNR_average_int)]] <- names(int_list)[[i]]
    }
    
    if(!any(grepl("NaN", hrIFNy))){
      hrIFNy_average_int <- rbind(hrIFNy_average_int, hrIFNy)
      rownames(hrIFNy_average_int)[[nrow(hrIFNy_average_int)]] <- names(int_list)[[i]]
    }
    
  }
  
  #Average across all tumours, but save a backup for other heatmaps
  colnames(unstim_average_int) <- colnames(TLS_023_014_data)
  unstim_average_int_tumour <- unstim_average_int
  unstim_average_int <- colMeans(unstim_average_int)
  
  colnames(aPD1_average_int) <- colnames(TLS_023_014_data)
  aPD1_average_int_tumour <- aPD1_average_int
  aPD1_average_int <- colMeans(aPD1_average_int)
  
  colnames(aPD1_aIFNR_average_int) <- colnames(TLS_023_014_data)
  aPD1_aIFNR_average_int_tumour <- aPD1_aIFNR_average_int
  aPD1_aIFNR_average_int <- colMeans(aPD1_aIFNR_average_int)
  
  colnames(hrIFNy_average_int) <- colnames(TLS_023_014_data)
  hrIFNy_average_int_tumour <- hrIFNy_average_int
  hrIFNy_average_int <- colMeans(hrIFNy_average_int)
  
  #and tumour
  tumour_list <- list(tumour_023_013_data,tumour_023_014_data, tumour_035_013_data, tumour_035_015_data, tumour_052_010_data, tumour_052_014_data,tumour_066_012_data,
                      tumour_066_015_data, tumour_080_data, tumour_083_012_data, tumour_083_015_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data)
  names(tumour_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012", "LU066_TLS015",
                       "LU080","LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
  unstim_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Calculate condition averages per tumour
  for(i in seq_along(tumour_list)){
    data <- tumour_list[[i]]
    
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    unstim <- colMeans(unstim)
    aPD1 <- colMeans(aPD1)
    aPD1_aIFNR <- colMeans(aPD1_aIFNR)
    hrIFNy <- colMeans(hrIFNy)
    
    if(!any(grepl("NaN", unstim))){
    unstim_average_tumour <- rbind(unstim_average_tumour, unstim)
    rownames(unstim_average_tumour)[[nrow(unstim_average_tumour)]] <- names(tumour_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1))){
      aPD1_average_tumour <- rbind(aPD1_average_tumour, aPD1)
      rownames(aPD1_average_tumour)[[nrow(aPD1_average_tumour)]] <- names(tumour_list)[[i]]
    }
    
    if(!any(grepl("NaN", aPD1_aIFNR))){
      aPD1_aIFNR_average_tumour <- rbind(aPD1_aIFNR_average_tumour, aPD1_aIFNR)
      rownames(aPD1_aIFNR_average_tumour)[[nrow(aPD1_aIFNR_average_tumour)]] <- names(tumour_list)[[i]]
    }
    
    if(!any(grepl("NaN", hrIFNy))){
      hrIFNy_average_tumour <- rbind(hrIFNy_average_tumour, hrIFNy)
      rownames(hrIFNy_average_tumour)[[nrow(hrIFNy_average_tumour)]] <- names(tumour_list)[[i]]
    }
    
  }
  
  #Average across all tumours, but save a backup for other heatmaps
  colnames(unstim_average_tumour) <- colnames(TLS_023_014_data)
  unstim_average_tumour_tumour <- unstim_average_tumour
  unstim_average_tumour <- colMeans(unstim_average_tumour)
  
  colnames(aPD1_average_tumour) <- colnames(TLS_023_014_data)
  aPD1_average_tumour_tumour <- aPD1_average_tumour
  aPD1_average_tumour <- colMeans(aPD1_average_tumour)
  
  colnames(aPD1_aIFNR_average_tumour) <- colnames(TLS_023_014_data)
  aPD1_aIFNR_average_tumour_tumour <- aPD1_aIFNR_average_tumour
  aPD1_aIFNR_average_tumour <- colMeans(aPD1_aIFNR_average_tumour)
  
  colnames(hrIFNy_average_tumour) <- colnames(TLS_023_014_data)
  hrIFNy_average_tumour_tumour <- hrIFNy_average_tumour
  hrIFNy_average_tumour <- colMeans(hrIFNy_average_tumour)
  
  #combine in a heatmap
  heatmap <- rbind(unstim_average_TLS,unstim_average_int,unstim_average_tumour,aPD1_average_TLS,aPD1_average_int,aPD1_average_tumour,
                   aPD1_aIFNR_average_TLS,aPD1_aIFNR_average_int,aPD1_aIFNR_average_tumour,hrIFNy_average_TLS,hrIFNy_average_int,hrIFNy_average_tumour)

  rownames(heatmap) <- c("unstim TLS" ,"unstim intermediate","unstim tumour bed","aPD1 TLS","aPD1 intermediate","aPD1 tumour bed",
                         "aPD1+aIFNyR1 TLS","aPD1+aIFNyR1 intermediate","aPD1+aIFNyR1 tumour bed","hrIFNy TLS","hrIFNy intermediate","hrIFNy tumour bed")  

  #z-scoring
  colSD <- apply(heatmap, 2, sd) #calculate standard deviation per cytokine/chemokine
  colMean <- colMeans(heatmap) #calculate means per cytokine/chemokine
  z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
  z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
  z_scores <- t(z_scores)

  rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y" #to make sure rownames can be seen
  rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"
  
  #annotation
  annotation <- data.frame(segment = rep(c("TLS", "intermediate", "tumor bed"),4), condition = rep(c("unstim","aPD1","aPD1+aIFNyR1","hrIFNy"), each = 3))
  rownames(annotation) <- colnames(z_scores)
  
  #heatmap
  heatmap <- pheatmap(z_scores, annotation_col = annotation,cluster_cols = F)
  
  save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_condition_heatmap_normalised_windsor.pdf") #save as PDF
```

Next, we generate a heatmap consisting of all individual fragments (TLS unstim LU023 fragment 1, TLS unstim LU023 fragment 2, etc.). For this, we use the data frames generated in the previous section, as they already contain all fragments split per segment type and condition.
```{r Fragment-based heatmap}
  #Per tumour; use the backups saved in the previous section
  rownames(unstim_average_TLS_tumour) <- paste0("unstim TLS ", rownames(unstim_average_TLS_tumour)) #name the fragment with its condition and segment type
  colnames(unstim_average_TLS_tumour) <- colnames(TLS_023_014_data) #add the cytokine/chemokine names
  rownames(aPD1_average_TLS_tumour) <- paste0("aPD1 TLS ", rownames(aPD1_average_TLS_tumour))
  colnames(aPD1_average_TLS_tumour) <- colnames(TLS_023_014_data)
  rownames(aPD1_aIFNR_average_TLS_tumour) <- paste0("aPD1+aIFNyR1 TLS ", rownames(aPD1_aIFNR_average_TLS_tumour))
  colnames(aPD1_aIFNR_average_TLS_tumour) <- colnames(TLS_023_014_data)
  rownames(hrIFNy_average_TLS_tumour) <- paste0("hrIFNy TLS ", rownames(hrIFNy_average_TLS_tumour))
  colnames(hrIFNy_average_TLS_tumour) <- colnames(TLS_023_014_data)
  
  #for intermediate
  rownames(unstim_average_int_tumour) <- paste0("unstim int ", rownames(unstim_average_int_tumour)) #name the fragment with its condition and segment type
  colnames(unstim_average_int_tumour) <- colnames(int_023_014_data) #add the cytokine/chemokine names
  rownames(aPD1_average_int_tumour) <- paste0("aPD1 int ", rownames(aPD1_average_int_tumour))
  colnames(aPD1_average_int_tumour) <- colnames(int_023_014_data)
  rownames(aPD1_aIFNR_average_int_tumour) <- paste0("aPD1+aIFNyR1 int ", rownames(aPD1_aIFNR_average_int_tumour))
  colnames(aPD1_aIFNR_average_int_tumour) <- colnames(int_023_014_data)
  rownames(hrIFNy_average_int_tumour) <- paste0("hrIFNy int ", rownames(hrIFNy_average_int_tumour))
  colnames(hrIFNy_average_int_tumour) <- colnames(int_023_014_data)
  
  #for tumour
  rownames(unstim_average_tumour_tumour) <- paste0("unstim tumour ", rownames(unstim_average_tumour_tumour)) #name the fragment with its condition and segment type
  colnames(unstim_average_tumour_tumour) <- colnames(tumour_023_014_data) #add the cytokine/chemokine names
  rownames(aPD1_average_tumour_tumour) <- paste0("aPD1 tumour ", rownames(aPD1_average_tumour_tumour))
  colnames(aPD1_average_tumour_tumour) <- colnames(tumour_023_014_data)
  rownames(aPD1_aIFNR_average_tumour_tumour) <- paste0("aPD1+aIFNyR1 tumour ", rownames(aPD1_aIFNR_average_tumour_tumour))
  colnames(aPD1_aIFNR_average_tumour_tumour) <- colnames(tumour_023_014_data)
  rownames(hrIFNy_average_tumour_tumour) <- paste0("hrIFNy tumour ", rownames(hrIFNy_average_tumour_tumour))
  colnames(hrIFNy_average_tumour_tumour) <- colnames(tumour_023_014_data)
  
  #combine in a heatmap
  heatmap <- rbind(unstim_per_tumour_TLS,unstim_per_tumour_int,unstim_per_tumour_tumour,aPD1_per_tumour_TLS,aPD1_per_tumour_int,aPD1_per_tumour_tumour,
                   aPD1_aIFNR_per_tumour_TLS,aPD1_aIFNR_per_tumour_int,aPD1_aIFNR_per_tumour_tumour,hrIFNy_per_tumour_TLS,hrIFNy_per_tumour_int,hrIFNy_per_tumour_tumour)
  
  #z-scoring
  colSD <- apply(heatmap, 2, sd)
  colMean <- colMeans(heatmap)
  z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
  z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
  z_scores <- t(z_scores)
  
  #heatmap
  rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y"
  rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"
  
  #a heatmap where the fragments are not clustered
  unclust <- pheatmap(z_scores,cluster_cols = F)
  save_pheatmap(unclust, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_unclust.pdf")
  save_pheatmap(unclust, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_unclust.png")
  
  #a version where the fragments are clustered
  clust <- pheatmap(z_scores,cluster_cols = T)
  save_pheatmap(unclust, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_clust.pdf")
  save_pheatmap(unclust, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_clust.png")
  
```

## 4. Data with outlier removal
Some fragments have very high scores for certain cytokines or chemokines. We do not expect such high values to be biological, rather, they are likely technical errors. To exclude such outliers from our data, we use the z-score: if the z-score is larger than 3 or smaller than -3 (the data point is more than 3 standard deviations away from the mean), we consider it an outlier and remove it. The fragment itself is not excluded, only the specific outlier value (for example, only the CXCL9 value of fragment Int LU083 aPD1 fragment 1). Z-scores are calculated from a data frame containing all fragments that we have.

```{r Outlier removal}
#find fragments with high z-scores
#since this just concerns a dataframe of all fragments, paste all original data together (no condition-based sorting needed):
lu023_TLS013_copy <- lu023_TLS013
rownames(lu023_TLS013_copy) <- gsub("LU023", "LU023_TLS013", rownames(lu023_TLS013_copy)) #to prevent duplicate fragment names, add the experiment number to the name

lu023_TLS014_copy <- lu023_TLS014
rownames(lu023_TLS014_copy) <- gsub("LU023", "LU023_TLS014", rownames(lu023_TLS014_copy)) #same here

lu035_TLS013_copy <- lu035_TLS013
rownames(lu035_TLS013_copy) <- gsub("TLS013 LU035", "LU035_TLS013", rownames(lu035_TLS013_copy))

lu035_TLS015_copy <- lu035_TLS015
rownames(lu035_TLS015_copy) <- gsub("TLS015 LU035", "LU035_TLS015", rownames(lu035_TLS015_copy))

lu052_TLS010_copy <- lu052_TLS010
rownames(lu052_TLS010_copy) <- gsub("TLS010 LU052", "LU052_TLS010", rownames(lu052_TLS010_copy))

lu052_TLS014_copy <- lu052_TLS014
rownames(lu052_TLS014_copy) <- gsub("TLS014 LU052", "LU052_TLS014", rownames(lu052_TLS014_copy))

lu066_TLS012_copy <- lu066_TLS012
rownames(lu066_TLS012_copy) <- gsub("LU066", "LU066_TLS012", rownames(lu066_TLS012_copy))

lu066_TLS015_copy <- lu066_TLS015
rownames(lu066_TLS015_copy) <- gsub("LU066", "LU066_TLS015", rownames(lu066_TLS015_copy))

lu066_TLS012_copy <- lu066_TLS012
rownames(lu066_TLS012_copy) <- gsub("LU066", "LU066_TLS012", rownames(lu066_TLS012_copy))

lu083_TLS012_copy <- lu083_TLS012
rownames(lu083_TLS012_copy) <- gsub("LU083", "LU083_TLS012", rownames(lu083_TLS012_copy))

lu083_TLS015_copy <- lu083_TLS015
rownames(lu083_TLS015_copy) <- gsub("LU083", "LU083_TLS015", rownames(lu083_TLS015_copy))

#combine all data together
full_df <- rbind(lu023_TLS013_copy, lu023_TLS014_copy, lu035_TLS013_copy, lu035_TLS015_copy, lu052_TLS010_copy, lu052_TLS014_copy, lu066_TLS012_copy, lu066_TLS015_copy,
                 lu080, lu083_TLS012_copy, lu083_TLS015_copy, lu086_1, lu086_2, lu088, lu091)

full_df <- full_df[!grepl("average|RN440|hrIFNy_high", rownames(full_df)),] #exclude averages, RN440 and hrIFNy_high fragments


#z-scoring
colSD <- apply(full_df, 2, sd)
colMean <- colMeans(full_df)
outliers <- sweep(full_df, 2, colMean, "-") #subtract means per column
outliers <- sweep(outliers, 2, colSD, "/") #divide by standard deviation per column
outliers <- t(outliers)

outliers[outliers > 3 | outliers < -3] <- NA #filter out datapoints (not the entire fragment) with individual high z-scores (above 3 or below -3)

#save outlier fragments in a separate data frame
outlier_fragment <- data.frame(matrix(nrow = 25, ncol = 0))

for(i in 1:ncol(outliers)){
  if(any(is.na(outliers[,i]))){ #if the fragment has an NA value, and therefore contains an outlier
    outlier_fragment <- cbind(outlier_fragment, outliers[,i]) #add to the data frame
    colnames(outlier_fragment)[[ncol(outlier_fragment)]] <- colnames(outliers)[i]
  } else {
    next
  }
}

#save result
write.csv(outlier_fragment, "~/Documents/Analyses/TLS/final_all_fragments/outliers.csv")
```

We can now re-do all of our previous analysis, but now with outlier values excluded. First, we generate individual heatmaps for each tumour and save these separately.

```{r Generate individual heatmaps for every tumour}
#redo heatmaps based on outlier information
#Get separate Legendplex scores per segment
TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,]
TLS_023_013_data <- check_outliers_fragment(TLS_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013") #the check_outlier function is there to set any outlier values found in the outliers data frame to NA
int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,]
int_023_013_data <- check_outliers_fragment(int_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")
tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,]
tumour_023_013_data <- check_outliers_fragment(tumour_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")

TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
TLS_023_014_data <- check_outliers_fragment(TLS_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
int_023_014_data <- check_outliers_fragment(int_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]
tumour_023_014_data <- check_outliers_fragment(tumour_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")

TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
TLS_035_013_data <- check_outliers_fragment(TLS_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
int_035_013_data <- check_outliers_fragment(int_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]
tumour_035_013_data <- check_outliers_fragment(tumour_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")

TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
TLS_035_015_data <- check_outliers_fragment(TLS_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
int_035_015_data <- check_outliers_fragment(int_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]
tumour_035_015_data <- check_outliers_fragment(tumour_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")

TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
TLS_052_010_data <- check_outliers_fragment(TLS_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
TLS_052_010_data <- TLS_052_010_data[!grepl("hrIFNy_high", rownames(TLS_052_010_data)),] #take out hrIFNy_high fragments
int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
int_052_010_data <- check_outliers_fragment(int_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
int_052_010_data <- int_052_010_data[!grepl("hrIFNy_high", rownames(int_052_010_data)),]
tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
tumour_052_010_data <- check_outliers_fragment(tumour_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
tumour_052_010_data <- tumour_052_010_data[!grepl("hrIFNy_high", rownames(tumour_052_010_data)),]

TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
TLS_052_014_data <- check_outliers_fragment(TLS_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
int_052_014_data <- check_outliers_fragment(int_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]
tumour_052_014_data <- check_outliers_fragment(tumour_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")

#This tumour has no intermediate fragments. Therefore, this condition is skipped.
TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
TLS_066_012_data <- check_outliers_fragment(TLS_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")
tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]
tumour_066_012_data <- check_outliers_fragment(tumour_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")

TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
TLS_066_015_data <- check_outliers_fragment(TLS_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
int_066_015_data <- check_outliers_fragment(int_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]
tumour_066_015_data <- check_outliers_fragment(tumour_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")

#LU080 has no TLS so we ignore this condition
int_080_data <- lu080[rownames(lu080) %in% int_080,]
int_080_data <- check_outliers_fragment(int_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)
#one cytokine (CXCL9) is NA/outlier for all fragments in this tumour, set to NA:
int_080_data[int_080_data == "NaN"] <- NA

tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]
tumour_080_data <- check_outliers_fragment(tumour_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)

TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
TLS_083_012_data <- check_outliers_fragment(TLS_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
int_083_012_data <- check_outliers_fragment(int_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]
tumour_083_012_data <- check_outliers_fragment(tumour_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")

TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
TLS_083_015_data <- check_outliers_fragment(TLS_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
int_083_015_data <- check_outliers_fragment(int_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]
tumour_083_015_data <- check_outliers_fragment(tumour_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")

#the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
int_086_1_data <- check_outliers_fragment(int_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)
tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]
tumour_086_1_data <- check_outliers_fragment(tumour_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)

#the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
int_086_2_data <- check_outliers_fragment(int_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)
tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]
tumour_086_2_data <- check_outliers_fragment(tumour_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)

TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
TLS_088_data <- check_outliers_fragment(TLS_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
int_088_data <- lu088[rownames(lu088) %in% int_088,]
int_088_data <- check_outliers_fragment(int_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]
tumour_088_data <- check_outliers_fragment(tumour_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)

TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
TLS_091_data <- check_outliers_fragment(TLS_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
int_091_data <- lu091[rownames(lu091) %in% int_091,]
int_091_data <- check_outliers_fragment(int_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]
tumour_091_data <- check_outliers_fragment(tumour_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)


#heatmap per tumour
generate_heatmap <- function(TLS_data, int_data, tumour_data, tumour_name, experiment){
  z_score <- rbind(TLS_data,int_data,tumour_data) #collect all outlier-removed data per tumour
  annotation <- data.frame(segment = c(rep("TLS", nrow(TLS_data)), rep("intermediate", nrow(int_data)), rep("tumour bed", nrow(tumour_data)))) #make the annotation based on the amount of fragments in each data frame
  rownames(annotation) <- rownames(z_score) #set back the fragment names
  
  #z-scoring
  colSD <- apply(z_score, 2, sd, na.rm = T)
  colSD[is.na(colSD)] <- 0 #happens when only one fragment has a value
  colMean <- colMeans(z_score, na.rm = T)
  z_score <- sweep(z_score, 2, colMean, "-") #subtract means per column
  z_score<- sweep(z_score, 2, colSD, "/") #divide by standard deviation per column
  z_score[z_score == "NaN"] <- 0 #happens when SD is zero
  z_score<- t(z_score)
  
  #one entry (LU080, CXCL9) has no values anywhere (except one fragment, but it makes no sense to calculate distances on one fragment): remove this mediator
  na_count <- rowSums(is.na(z_score)) #count the number of NA values in each row (mediator)
  z_score <- z_score[na_count < (ncol(z_score)-1),] #only keep the mediator if the amount of NAs is lower than the amount of samples -1 (2 or more non-NA values should exist)
  
  #make a heatmap
  heatmap<- pheatmap(z_score, annotation_col = annotation, cluster_cols = F)
  
  #store the result
  pdf(paste0("~/Documents/Analyses/TLS/final_all_fragments/heatmap_", tumour_name, "_", experiment, ".pdf"),width = 15)
  heatmap
  dev.off()
  
  save_pheatmap(heatmap, paste0("~/Documents/Analyses/TLS/final_all_fragments/heatmap_", tumour_name, "_", experiment, ".png"))
}

#generate heatmaps for every tumour
generate_heatmap(TLS_023_013_data,int_023_013_data,tumour_023_013_data, "LU023", "TLS013")
dev.off()
generate_heatmap(TLS_023_014_data,int_023_014_data,tumour_023_014_data, "LU023", "TLS014")
dev.off()
generate_heatmap(TLS_035_013_data,int_035_013_data,tumour_035_013_data, "LU035", "TLS013")
dev.off()
generate_heatmap(TLS_035_015_data,int_035_015_data,tumour_035_015_data, "LU035", "TLS015")
dev.off()
generate_heatmap(TLS_052_010_data,int_052_010_data,tumour_052_010_data, "LU052", "TLS010")
dev.off()
generate_heatmap(TLS_052_014_data,int_052_014_data,tumour_052_014_data, "LU052", "TLS014")
dev.off()
generate_heatmap(TLS_066_012_data,data.frame(),tumour_066_012_data, "LU066", "TLS012") #this entry has no intermediate fragments
dev.off()
generate_heatmap(TLS_066_015_data,int_066_015_data,tumour_066_015_data, "LU066", "TLS015")
dev.off()
generate_heatmap(TLS_data = data.frame(), int_data = int_080_data, tumour_data = tumour_080_data, tumour =  "LU080", experiment = "TLS011") #this entry has no TLS or CXCL9 (all outlier)
dev.off()
generate_heatmap(TLS_083_012_data,int_083_012_data,tumour_083_012_data, "LU083", "TLS012")
dev.off()
generate_heatmap(TLS_083_015_data,int_083_015_data,tumour_083_015_data, "LU083", "TLS015")
dev.off()
generate_heatmap(TLS_data = data.frame(),int_086_1_data,tumour_086_1_data, "LU086-1", "TLS011") #this entry has no TLS
dev.off()
generate_heatmap(TLS_data = data.frame(),int_086_2_data,tumour_086_2_data, "LU086-2", "TLS011") #this entry has no TLS
dev.off()
generate_heatmap(TLS_088_data,int_088_data,tumour_088_data, "LU088", "TLS016")
dev.off()
generate_heatmap(TLS_091_data,int_091_data,tumour_091_data, "LU091", "TLS016")
dev.off()
```

Next, we generate a heatmap where we take the average per tumour, segment type and condition (TLS LU023 unstim, Int LU023 unstim, .... TLS LU023 aPD1 and so on).
```{r Segment/condition/tumour based heatmap}
#calculate averages for all tumours
TLS_023_013_data <- colMeans(TLS_023_013_data, na.rm = T) #average score per cytokine across all conditions (ignore NA values)
int_023_013_data <- colMeans(int_023_013_data, na.rm = T)
tumour_023_013_data <- colMeans(tumour_023_013_data, na.rm = T)
TLS_023_014_data <- colMeans(TLS_023_014_data, na.rm = T) #average score per cytokine across all conditions
int_023_014_data <- colMeans(int_023_014_data, na.rm = T)
tumour_023_014_data <- colMeans(tumour_023_014_data, na.rm = T)
TLS_035_013_data <- colMeans(TLS_035_013_data, na.rm = T) #average score per cytokine across all conditions
int_035_013_data <- colMeans(int_035_013_data, na.rm = T)
tumour_035_013_data <- colMeans(tumour_035_013_data, na.rm = T)
TLS_035_015_data <- colMeans(TLS_035_015_data, na.rm = T) #average score per cytokine across all conditions
int_035_015_data <- colMeans(int_035_015_data, na.rm = T)
tumour_035_015_data <- colMeans(tumour_035_015_data, na.rm = T)
TLS_052_010_data <- colMeans(TLS_052_010_data, na.rm = T) #average score per cytokine across all conditions
int_052_010_data <- colMeans(int_052_010_data, na.rm = T)
tumour_052_010_data <- colMeans(tumour_052_010_data, na.rm = T)
TLS_052_014_data <- colMeans(TLS_052_014_data, na.rm = T) #average score per cytokine across all conditions
int_052_014_data <- colMeans(int_052_014_data, na.rm = T)
tumour_052_014_data <- colMeans(tumour_052_014_data, na.rm = T)
TLS_066_012_data <- colMeans(TLS_066_012_data, na.rm = T) #average score per cytokine across all conditions
tumour_066_012_data <- colMeans(tumour_066_012_data, na.rm = T)
TLS_066_015_data <- colMeans(TLS_066_015_data, na.rm = T) #average score per cytokine across all conditions
int_066_015_data <- colMeans(int_066_015_data, na.rm = T)
tumour_066_015_data <- colMeans(tumour_066_015_data, na.rm = T)
int_080_data <- colMeans(int_080_data, na.rm = T)
tumour_080_data <- colMeans(tumour_080_data, na.rm = T)
TLS_083_012_data <- colMeans(TLS_083_012_data, na.rm = T) #average score per cytokine across all conditions
int_083_012_data <- colMeans(int_083_012_data, na.rm = T)
tumour_083_012_data <- colMeans(tumour_083_012_data, na.rm = T)
TLS_083_015_data <- colMeans(TLS_083_015_data, na.rm = T) #average score per cytokine across all conditions
int_083_015_data <- colMeans(int_083_015_data, na.rm = T)
tumour_083_015_data <- colMeans(tumour_083_015_data, na.rm = T)
int_086_1_data <- colMeans(int_086_1_data, na.rm = T)
tumour_086_1_data <- colMeans(tumour_086_1_data, na.rm = T)
int_086_2_data <- colMeans(int_086_2_data, na.rm = T)
tumour_086_2_data <- colMeans(tumour_086_2_data, na.rm = T)
TLS_088_data <- colMeans(TLS_088_data, na.rm = T) #average score per cytokine across all conditions
int_088_data <- colMeans(int_088_data, na.rm = T)
tumour_088_data <- colMeans(tumour_088_data, na.rm = T)
TLS_091_data <- colMeans(TLS_091_data, na.rm = T) #average score per cytokine across all conditions
int_091_data <- colMeans(int_091_data, na.rm = T)
tumour_091_data <- colMeans(tumour_091_data, na.rm = T)

#generate heatmaps per tumour and segment
#Combine data from all tumours
TLS_scores <- rbind(TLS_023_013_data,TLS_023_014_data, TLS_035_013_data, TLS_035_015_data, TLS_052_010_data, TLS_052_014_data, TLS_066_012_data,TLS_066_015_data,
                    TLS_083_012_data, TLS_083_015_data, TLS_088_data, TLS_091_data) #missing LU080, LU086-1, LU086-2
TLS_scores_mean <- colMeans(TLS_scores, na.rm = T) #save mean across all tumours for later
rownames(TLS_scores) <- gsub("TLS_", "TLS LU", rownames(TLS_scores)) #change observation names
rownames(TLS_scores) <- gsub("_data", "", rownames(TLS_scores))
rownames(TLS_scores) <- gsub("_0", " TLS0", rownames(TLS_scores))

#repeat for intermediate and tumour fragments
int_scores <- rbind(int_023_013_data,int_023_014_data, int_035_013_data, int_035_015_data, int_052_010_data, int_052_014_data, int_066_015_data,
                    int_080_data, int_083_012_data, int_083_015_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data) #missing LU066 (TLS012)
int_scores_mean <- colMeans(int_scores, na.rm = T)
rownames(int_scores) <- gsub("int_", "Intermediate LU", rownames(int_scores))
rownames(int_scores) <- gsub("_data", "", rownames(int_scores))
rownames(int_scores) <- gsub("_0", " TLS0", rownames(int_scores))

tumour_scores <- rbind(tumour_023_013_data,tumour_023_014_data, tumour_035_013_data, tumour_035_015_data, tumour_052_010_data, tumour_052_014_data, tumour_066_012_data,tumour_066_015_data,
                       tumour_080_data, tumour_083_012_data, tumour_083_015_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data)
tumour_scores_mean <- colMeans(tumour_scores, na.rm = T)
rownames(tumour_scores) <- gsub("tumour_", "Tumour_bed LU", rownames(tumour_scores))
rownames(tumour_scores) <- gsub("_data", "", rownames(tumour_scores))
rownames(tumour_scores) <- gsub("_0", " TLS0", rownames(tumour_scores))

#make a heatmap of all tumour values
heatmap_scores_tumour <- rbind(TLS_scores, int_scores, tumour_scores) #combine fragments from all regions
#z-scoring
colSD <- apply(heatmap_scores_tumour, 2, sd, na.rm = T)
colMean <- colMeans(heatmap_scores_tumour, na.rm = T)
z_scores_tumour <- sweep(heatmap_scores_tumour, 2, colMean, "-") #subtract means per column
z_scores_tumour <- sweep(z_scores_tumour, 2, colSD, "/") #divide by standard deviation per column
z_scores_tumour <- t(z_scores_tumour)
rownames(z_scores_tumour)[rownames(z_scores_tumour) == "IFN-γ"] <- "IFN-y"
rownames(z_scores_tumour)[rownames(z_scores_tumour) == "TNF-α"] <- "TNF-a"

#add response score
annotation_tumour <- data.frame(PDTF_responder = rep(c("PDTF_NR","PDTF_R","PDTF_NR","None","PDTF_R","None","PDTF_NR","PDTF_NR","PDTF_NR","PDTF_R",
                                                        "PDTF_R","PDTF_NR","PDTF_NR","PDTF_NR","PDTF_NR"), times = 3)) #the order of this is based on the order of the tumours in each of the three data frames
#remove for segments that do not exist (LU066_012 int,LU080 TLS, etc.)
annotation_tumour <- as.data.frame(annotation_tumour[!(rownames(annotation_tumour) %in% c(22,12,13,9)),])
colnames(annotation_tumour) <- "PDTF_score"
rownames(annotation_tumour) <- colnames(z_scores_tumour)
annotation_tumour$PDTF_score <- factor(annotation_tumour$PDTF_score, levels = c("PDTF_NR", "PDTF_R", "None"))

#make the colours for the PDTF response score
ann_colors = list(
  PDTF_score = c(PDTF_NR = "pink",PDTF_R= "lightgreen",None= "grey"))

#make and save unclustered and clustered heatmaps (clustering on the fragment level)
heatmap <- suppressWarnings(pheatmap(z_scores_tumour, annotation_col = annotation_tumour, annotation_colors = ann_colors))  
heatmap
save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved.pdf")
save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved.png")

heatmap_unclust <- suppressWarnings(pheatmap(z_scores_tumour, annotation_col = annotation_tumour, annotation_colors = ann_colors, cluster_cols = F))  
save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_unclust.pdf")
save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_tumours_normalised_outlierremoved_unclust.png")
```

Next, we generate a heatmap for the segment types (TLS vs Int vs tumour bed), averaged for all tumours and conditions.
```{r Segment based heatmap}
#Generate a heatmap for all averaged segments (so one column for TLS, intermediate and low)
heatmap_scores <- rbind(TLS_scores_mean, int_scores_mean, tumour_scores_mean) #combine the means calculated in the previous section
rownames(heatmap_scores) <- c("TLS", "intermediate", "tumour bed")
colSD <- apply(heatmap_scores, 2, sd)
colMean <- colMeans(heatmap_scores)
z_scores <- sweep(heatmap_scores, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)
rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y"
rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"

#Make and save the heatmap
heatmap <- suppressWarnings(pheatmap(z_scores))  
heatmap

save_pheatmap(heatmap, "~/Documents/Analyses/TLS/final_all_fragments/segment_normalised_outlierremoved.pdf")
```

We now generate a heatmap per segemnt type and conditon, averaged across all tumours (TLS unstim, TLS aPD1... Int unstim, Int aPD1 etc.)
```{r Segment/condition based heatmap}
#Make a heatmap of condition averages per fragment type (TLS-unstim, TLS-aPD1, etc.)
TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,]
TLS_023_013_data <- check_outliers_fragment(TLS_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013") #function to remove outlier values
int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,]
int_023_013_data <- check_outliers_fragment(int_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")
tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,]
tumour_023_013_data <- check_outliers_fragment(tumour_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")

TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
TLS_023_014_data <- check_outliers_fragment(TLS_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
int_023_014_data <- check_outliers_fragment(int_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]
tumour_023_014_data <- check_outliers_fragment(tumour_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")

TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
TLS_035_013_data <- check_outliers_fragment(TLS_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
int_035_013_data <- check_outliers_fragment(int_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]
tumour_035_013_data <- check_outliers_fragment(tumour_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")

TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
TLS_035_015_data <- check_outliers_fragment(TLS_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
int_035_015_data <- check_outliers_fragment(int_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]
tumour_035_015_data <- check_outliers_fragment(tumour_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")

TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
TLS_052_010_data <- check_outliers_fragment(TLS_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
int_052_010_data <- check_outliers_fragment(int_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
tumour_052_010_data <- check_outliers_fragment(tumour_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
#LU052_TLS010_tumour_bed has extremely high CXCL9/CXCL10, what does the heatmap look like if I take the IFN_high fragments out?
TLS_052_010_data <- TLS_052_010_data[!grepl("hrIFNy_high", rownames(TLS_052_010_data)),]
int_052_010_data <- int_052_010_data[!grepl("hrIFNy_high", rownames(int_052_010_data)),]
tumour_052_010_data <- tumour_052_010_data[!grepl("hrIFNy_high", rownames(tumour_052_010_data)),]

TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
TLS_052_014_data <- check_outliers_fragment(TLS_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
int_052_014_data <- check_outliers_fragment(int_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]
tumour_052_014_data <- check_outliers_fragment(tumour_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")

#This tumour has no intermediate fragments. Therefore, this condition is skipped.
TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
TLS_066_012_data <- check_outliers_fragment(TLS_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")
tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]
tumour_066_012_data <- check_outliers_fragment(tumour_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")

TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
TLS_066_015_data <- check_outliers_fragment(TLS_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
int_066_015_data <- check_outliers_fragment(int_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]
tumour_066_015_data <- check_outliers_fragment(tumour_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")

#LU080 has no TLS so we ignore this condition
int_080_data <- lu080[rownames(lu080) %in% int_080,]
int_080_data <- check_outliers_fragment(int_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)
tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]
tumour_080_data <- check_outliers_fragment(tumour_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)

TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
TLS_083_012_data <- check_outliers_fragment(TLS_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
int_083_012_data <- check_outliers_fragment(int_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]
tumour_083_012_data <- check_outliers_fragment(tumour_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")

TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
TLS_083_015_data <- check_outliers_fragment(TLS_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
int_083_015_data <- check_outliers_fragment(int_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]
tumour_083_015_data <- check_outliers_fragment(tumour_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")

#the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
int_086_1_data <- check_outliers_fragment(int_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)
tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]
tumour_086_1_data <- check_outliers_fragment(tumour_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)

#the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
int_086_2_data <- check_outliers_fragment(int_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)
tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]
tumour_086_2_data <- check_outliers_fragment(tumour_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)

TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
TLS_088_data <- check_outliers_fragment(TLS_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
int_088_data <- lu088[rownames(lu088) %in% int_088,]
int_088_data <- check_outliers_fragment(int_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]
tumour_088_data <- check_outliers_fragment(tumour_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)

TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
TLS_091_data <- check_outliers_fragment(TLS_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
int_091_data <- lu091[rownames(lu091) %in% int_091,]
int_091_data <- check_outliers_fragment(int_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]
tumour_091_data <- check_outliers_fragment(tumour_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)

#First for TLS
TLS_list <- list(TLS_023_013_data,TLS_023_014_data, TLS_035_013_data, TLS_035_015_data, TLS_052_010_data, TLS_052_014_data, TLS_066_012_data,TLS_066_015_data,
                 TLS_083_012_data, TLS_083_015_data, TLS_088_data, TLS_091_data) #missing LU080, LU086-1, LU086-2
names(TLS_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012","LU066_TLS015",
                     "LU083_TLS012", "LU083_TLS015", "LU088", "LU091")  

#create data frames for each condition
unstim_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(TLS_list)){ #for TLS segments of each tumour
  data <- TLS_list[[i]] #grab the data
  
  #split the fragments across the different conditions
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  #calculate the tumour average for each condition
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  #if the tumour has the condition, add it to its respective data frame
  if(!any(grepl("NaN", unstim))){
    unstim_average_TLS <- rbind(unstim_average_TLS, unstim)
    rownames(unstim_average_TLS)[[nrow(unstim_average_TLS)]] <- names(TLS_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_TLS <- rbind(aPD1_average_TLS, aPD1)
    rownames(aPD1_average_TLS)[[nrow(aPD1_average_TLS)]] <- names(TLS_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_TLS <- rbind(aPD1_aIFNR_average_TLS, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_TLS)[[nrow(aPD1_aIFNR_average_TLS)]] <- names(TLS_list)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_TLS <- rbind(hrIFNy_average_TLS, hrIFNy)
    rownames(hrIFNy_average_TLS)[[nrow(hrIFNy_average_TLS)]] <- names(TLS_list)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_TLS) <- colnames(TLS_023_014_data)
unstim_average_TLS <- colMeans(unstim_average_TLS, na.rm = T)

colnames(aPD1_average_TLS) <- colnames(TLS_023_014_data)
aPD1_average_TLS <- colMeans(aPD1_average_TLS, na.rm = T)

colnames(aPD1_aIFNR_average_TLS) <- colnames(TLS_023_014_data)
aPD1_aIFNR_average_TLS <- colMeans(aPD1_aIFNR_average_TLS, na.rm = T)

colnames(hrIFNy_average_TLS) <- colnames(TLS_023_014_data)
hrIFNy_average_TLS <- colMeans(hrIFNy_average_TLS, na.rm = T)

#repeat for intermediate and tumour sections
int_list <- list(int_023_013_data,int_023_014_data, int_035_013_data, int_035_015_data, int_052_010_data, int_052_014_data, int_066_015_data,
                 int_080_data, int_083_012_data, int_083_015_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data) #missing LU066 (TLS012)
names(int_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014", "LU066_TLS015",
                     "LU080","LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
unstim_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_int <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(int_list)){
  data <- int_list[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_int <- rbind(unstim_average_int, unstim)
    rownames(unstim_average_int)[[nrow(unstim_average_int)]] <- names(int_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_int <- rbind(aPD1_average_int, aPD1)
    rownames(aPD1_average_int)[[nrow(aPD1_average_int)]] <- names(int_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_int <- rbind(aPD1_aIFNR_average_int, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_int)[[nrow(aPD1_aIFNR_average_int)]] <- names(int_list)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_int <- rbind(hrIFNy_average_int, hrIFNy)
    rownames(hrIFNy_average_int)[[nrow(hrIFNy_average_int)]] <- names(int_list)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_int) <- colnames(TLS_023_014_data)
unstim_average_int <- colMeans(unstim_average_int, na.rm = T)

colnames(aPD1_average_int) <- colnames(TLS_023_014_data)
aPD1_average_int <- colMeans(aPD1_average_int, na.rm = T)

colnames(aPD1_aIFNR_average_int) <- colnames(TLS_023_014_data)
aPD1_aIFNR_average_int <- colMeans(aPD1_aIFNR_average_int, na.rm = T)

colnames(hrIFNy_average_int) <- colnames(TLS_023_014_data)
hrIFNy_average_int <- colMeans(hrIFNy_average_int, na.rm = T)

#and tumour
tumour_list <- list(tumour_023_013_data,tumour_023_014_data, tumour_035_013_data, tumour_035_015_data, tumour_052_010_data, tumour_052_014_data,tumour_066_012_data,
                    tumour_066_015_data, tumour_080_data, tumour_083_012_data, tumour_083_015_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data)
names(tumour_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012", "LU066_TLS015",
                        "LU080","LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
unstim_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(tumour_list)){
  data <- tumour_list[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_tumour <- rbind(unstim_average_tumour, unstim)
    rownames(unstim_average_tumour)[[nrow(unstim_average_tumour)]] <- names(tumour_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_tumour <- rbind(aPD1_average_tumour, aPD1)
    rownames(aPD1_average_tumour)[[nrow(aPD1_average_tumour)]] <- names(tumour_list)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_tumour <- rbind(aPD1_aIFNR_average_tumour, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_tumour)[[nrow(aPD1_aIFNR_average_tumour)]] <- names(tumour_list)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_tumour <- rbind(hrIFNy_average_tumour, hrIFNy)
    rownames(hrIFNy_average_tumour)[[nrow(hrIFNy_average_tumour)]] <- names(tumour_list)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_tumour) <- colnames(TLS_023_014_data)
unstim_average_tumour <- colMeans(unstim_average_tumour, na.rm = T)

colnames(aPD1_average_tumour) <- colnames(TLS_023_014_data)
aPD1_average_tumour <- colMeans(aPD1_average_tumour, na.rm = T)

colnames(aPD1_aIFNR_average_tumour) <- colnames(TLS_023_014_data)
aPD1_aIFNR_average_tumour <- colMeans(aPD1_aIFNR_average_tumour, na.rm = T)

colnames(hrIFNy_average_tumour) <- colnames(TLS_023_014_data)
hrIFNy_average_tumour <- colMeans(hrIFNy_average_tumour, na.rm = T)




#combine in a heatmap
heatmap <- rbind(unstim_average_TLS,unstim_average_int,unstim_average_tumour,aPD1_average_TLS,aPD1_average_int,aPD1_average_tumour,
                 aPD1_aIFNR_average_TLS,aPD1_aIFNR_average_int,aPD1_aIFNR_average_tumour,hrIFNy_average_TLS,hrIFNy_average_int,hrIFNy_average_tumour)

rownames(heatmap) <- c("unstim TLS" ,"unstim intermediate","unstim tumour bed","aPD1 TLS","aPD1 intermediate","aPD1 tumour bed",
                       "aPD1+aIFNyR1 TLS","aPD1+aIFNyR1 intermediate","aPD1+aIFNyR1 tumour bed","hrIFNy TLS","hrIFNy intermediate","hrIFNy tumour bed")  

#z-scoring
colSD <- apply(heatmap, 2, sd)
colMean <- colMeans(heatmap)
z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y" #to make sure rownames can be seen
rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"

#annotation
annotation <- data.frame(segment = rep(c("TLS", "intermediate", "tumor bed"),4), condition = rep(c("unstim","aPD1","aPD1+aIFNyR1","hrIFNy"), each = 3))
rownames(annotation) <- colnames(z_scores)

#heatmap
correlations <- pheatmap(z_scores, annotation_col = annotation,cluster_cols = F)

correlations

#capture row order
row_order <- correlations$tree_row$order
values <- rownames(z_scores)[row_order] #grab the names of the cytokines in order of the heatmap

#reorder z-scores based on row (cytokine) order
z_scores <- z_scores[order(match(rownames(z_scores), values)),]

#save z-scores
z_scores <- as.data.frame(z_scores)
write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/z_scores_condition_segment_normalised_outlierremoved.xlsx", rowNames = T)

```

## 5. PDTF response score
Several of our tumours are PDTF responders (LU023, LU052 and LU083) while the others are PDTF non-responders. These likely have differing mediator patterns in the different segment and conditions. Therefore, we split our next analyses into PDTF-R and PDTF-NR.

In this next section, we generate three heatmaps: unstim-aPD1-aPD1+aIFNyR1 (PDTF-R only) and unstim-hrIFNy (PDTF-R and PDTF-NR), as it only makes sense to look at aPD1 response in PDTF-R tumours.
```{r Heatmaps for PDTF-R and PDTF-NR separately}
#split up between R and NR (unstim-aPD1-combi)
outlier_fragment <- read.csv("~/Documents/Analyses/TLS/final_all_fragments/outliers.csv") #grab the data frame containing the outlier fragment information

#grab TLS< intermediate and tumour fragments for each tumour
TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,]
TLS_023_013_data <- check_outliers_fragment(TLS_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013") #remove outliers
int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,]
int_023_013_data <- check_outliers_fragment(int_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")
tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,]
tumour_023_013_data <- check_outliers_fragment(tumour_023_013_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS013")

TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
TLS_023_014_data <- check_outliers_fragment(TLS_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
int_023_014_data <- check_outliers_fragment(int_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")
tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]
tumour_023_014_data <- check_outliers_fragment(tumour_023_014_data, outliers = outlier_fragment, tumour = "LU023", experiment = "TLS014")

TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
TLS_035_013_data <- check_outliers_fragment(TLS_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
int_035_013_data <- check_outliers_fragment(int_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")
tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]
tumour_035_013_data <- check_outliers_fragment(tumour_035_013_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS013")

TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
TLS_035_015_data <- check_outliers_fragment(TLS_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
int_035_015_data <- check_outliers_fragment(int_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")
tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]
tumour_035_015_data <- check_outliers_fragment(tumour_035_015_data, outliers = outlier_fragment, tumour = "LU035", experiment = "TLS015")

TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
TLS_052_010_data <- check_outliers_fragment(TLS_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
int_052_010_data <- check_outliers_fragment(int_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
tumour_052_010_data <- check_outliers_fragment(tumour_052_010_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS010")
#LU052_TLS010_tumour_bed has extremely high CXCL9/CXCL10, what does the heatmap look like if I take the IFN_high fragments out?
TLS_052_010_data <- TLS_052_010_data[!grepl("hrIFNy_high", rownames(TLS_052_010_data)),]
int_052_010_data <- int_052_010_data[!grepl("hrIFNy_high", rownames(int_052_010_data)),]
tumour_052_010_data <- tumour_052_010_data[!grepl("hrIFNy_high", rownames(tumour_052_010_data)),]

TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
TLS_052_014_data <- check_outliers_fragment(TLS_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
int_052_014_data <- check_outliers_fragment(int_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")
tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]
tumour_052_014_data <- check_outliers_fragment(tumour_052_014_data, outliers = outlier_fragment, tumour = "LU052", experiment = "TLS014")

#This tumour has no intermediate fragments. Therefore, this condition is skipped.
TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
TLS_066_012_data <- check_outliers_fragment(TLS_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")
tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]
tumour_066_012_data <- check_outliers_fragment(tumour_066_012_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS012")

TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
TLS_066_015_data <- check_outliers_fragment(TLS_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
int_066_015_data <- check_outliers_fragment(int_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")
tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]
tumour_066_015_data <- check_outliers_fragment(tumour_066_015_data, outliers = outlier_fragment, tumour = "LU066", experiment = "TLS015")

#LU080 has no TLS so we ignore this condition
int_080_data <- lu080[rownames(lu080) %in% int_080,]
int_080_data <- check_outliers_fragment(int_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)
tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]
tumour_080_data <- check_outliers_fragment(tumour_080_data, outliers = outlier_fragment, tumour = "LU080", experiment = NA)

TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
TLS_083_012_data <- check_outliers_fragment(TLS_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
int_083_012_data <- check_outliers_fragment(int_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")
tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]
tumour_083_012_data <- check_outliers_fragment(tumour_083_012_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS012")

TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
TLS_083_015_data <- check_outliers_fragment(TLS_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
int_083_015_data <- check_outliers_fragment(int_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")
tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]
tumour_083_015_data <- check_outliers_fragment(tumour_083_015_data, outliers = outlier_fragment, tumour = "LU083", experiment = "TLS015")

#the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
int_086_1_data <- check_outliers_fragment(int_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)
tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]
tumour_086_1_data <- check_outliers_fragment(tumour_086_1_data, outliers = outlier_fragment, tumour = "LU086-1", experiment = NA)

#the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
int_086_2_data <- check_outliers_fragment(int_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)
tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]
tumour_086_2_data <- check_outliers_fragment(tumour_086_2_data, outliers = outlier_fragment, tumour = "LU086-2", experiment = NA)

TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
TLS_088_data <- check_outliers_fragment(TLS_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
int_088_data <- lu088[rownames(lu088) %in% int_088,]
int_088_data <- check_outliers_fragment(int_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)
tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]
tumour_088_data <- check_outliers_fragment(tumour_088_data, outliers = outlier_fragment, tumour = "LU088", experiment = NA)

TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
TLS_091_data <- check_outliers_fragment(TLS_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
int_091_data <- lu091[rownames(lu091) %in% int_091,]
int_091_data <- check_outliers_fragment(int_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)
tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]
tumour_091_data <- check_outliers_fragment(tumour_091_data, outliers = outlier_fragment, tumour = "LU091", experiment = NA)

#First for TLS: split into responders (LU023, LU052, LU083) and nonresponders (LU035, LU066, LU080, LU086, LU088, LU091)
TLS_list_R <- list(TLS_023_013_data,TLS_023_014_data, TLS_052_010_data, TLS_052_014_data,TLS_083_012_data, TLS_083_015_data)
names(TLS_list_R) <- c("LU023_TLS013","LU023_TLS014","LU052_TLS010","LU052_TLS014","LU083_TLS012", "LU083_TLS015")  

TLS_list_NR <- list(TLS_035_013_data, TLS_035_015_data, TLS_066_012_data,TLS_066_015_data, TLS_088_data, TLS_091_data)
names(TLS_list_NR) <- c("LU035_TLS013","LU035_TLS015","LU066_TLS012","LU066_TLS015","LU088", "LU091")  

#again; generate conditional averages per tumour, but now do this for responders and non-responders separately
unstim_average_TLS_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_TLS_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_TLS_R <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_TLS_R <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(TLS_list_R)){
  data <- TLS_list_R[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_TLS_R <- rbind(unstim_average_TLS_R, unstim)
    rownames(unstim_average_TLS_R)[[nrow(unstim_average_TLS_R)]] <- names(TLS_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_TLS_R <- rbind(aPD1_average_TLS_R, aPD1)
    rownames(aPD1_average_TLS_R)[[nrow(aPD1_average_TLS_R)]] <- names(TLS_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_TLS_R <- rbind(aPD1_aIFNR_average_TLS_R, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_TLS_R)[[nrow(aPD1_aIFNR_average_TLS_R)]] <- names(TLS_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_TLS_R <- rbind(hrIFNy_average_TLS_R, hrIFNy)
    rownames(hrIFNy_average_TLS_R)[[nrow(hrIFNy_average_TLS_R)]] <- names(TLS_list_R)[[i]]
  }
  
}

#Average across all responder tumours
colnames(unstim_average_TLS_R) <- colnames(TLS_023_014_data)
unstim_average_TLS_R <- colMeans(unstim_average_TLS_R, na.rm = T)

colnames(aPD1_average_TLS_R) <- colnames(TLS_023_014_data)
aPD1_average_TLS_R <- colMeans(aPD1_average_TLS_R, na.rm = T)

colnames(aPD1_aIFNR_average_TLS_R) <- colnames(TLS_023_014_data)
aPD1_aIFNR_average_TLS_R <- colMeans(aPD1_aIFNR_average_TLS_R, na.rm = T)

colnames(hrIFNy_average_TLS_R) <- colnames(TLS_023_014_data)
hrIFNy_average_TLS_R <- colMeans(hrIFNy_average_TLS_R, na.rm = T)

#repeat for non-responders
unstim_average_TLS_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_TLS_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_TLS_NR <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_TLS_NR <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(TLS_list_NR)){
  data <- TLS_list_NR[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_TLS_NR <- rbind(unstim_average_TLS_NR, unstim)
    rownames(unstim_average_TLS_NR)[[nrow(unstim_average_TLS_NR)]] <- names(TLS_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_TLS_NR <- rbind(aPD1_average_TLS_NR, aPD1)
    rownames(aPD1_average_TLS_NR)[[nrow(aPD1_average_TLS_NR)]] <- names(TLS_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_TLS_NR <- rbind(aPD1_aIFNR_average_TLS_NR, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_TLS_NR)[[nrow(aPD1_aIFNR_average_TLS_NR)]] <- names(TLS_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_TLS_NR <- rbind(hrIFNy_average_TLS_NR, hrIFNy)
    rownames(hrIFNy_average_TLS_NR)[[nrow(hrIFNy_average_TLS_NR)]] <- names(TLS_list_NR)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_TLS_NR) <- colnames(TLS_023_014_data)
unstim_average_TLS_NR <- colMeans(unstim_average_TLS_NR, na.rm = T)

colnames(aPD1_average_TLS_NR) <- colnames(TLS_023_014_data)
aPD1_average_TLS_NR <- colMeans(aPD1_average_TLS_NR, na.rm = T)

colnames(aPD1_aIFNR_average_TLS_NR) <- colnames(TLS_023_014_data)
aPD1_aIFNR_average_TLS_NR <- colMeans(aPD1_aIFNR_average_TLS_NR, na.rm = T)

colnames(hrIFNy_average_TLS_NR) <- colnames(TLS_023_014_data)
hrIFNy_average_TLS_NR <- colMeans(hrIFNy_average_TLS_NR, na.rm = T)



#repeat for intermediate and tumour sections
int_list_R <- list(int_023_013_data,int_023_014_data,int_052_010_data, int_052_014_data,int_083_012_data, int_083_015_data)
names(int_list_R) <- c("LU023_TLS013","LU023_TLS014","LU052_TLS010","LU052_TLS014", "LU083_TLS012", "LU083_TLS015")  

int_list_NR <- list(int_035_013_data, int_035_015_data, int_066_015_data,int_080_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data)
names(int_list_NR) <- c("LU035_TLS013","LU035_TLS015","LU066_TLS015","LU080", "LU086-1", "LU086-2", "LU088", "LU091")  

unstim_average_int_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_int_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_int_R <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_int_R <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(int_list_R)){
  data <- int_list_R[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_int_R <- rbind(unstim_average_int_R, unstim)
    rownames(unstim_average_int_R)[[nrow(unstim_average_int_R)]] <- names(int_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_int_R <- rbind(aPD1_average_int_R, aPD1)
    rownames(aPD1_average_int_R)[[nrow(aPD1_average_int_R)]] <- names(int_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_int_R <- rbind(aPD1_aIFNR_average_int_R, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_int_R)[[nrow(aPD1_aIFNR_average_int_R)]] <- names(int_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_int_R <- rbind(hrIFNy_average_int_R, hrIFNy)
    rownames(hrIFNy_average_int_R)[[nrow(hrIFNy_average_int_R)]] <- names(int_list_R)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_int_R) <- colnames(int_023_014_data)
unstim_average_int_R <- colMeans(unstim_average_int_R, na.rm = T)

colnames(aPD1_average_int_R) <- colnames(int_023_014_data)
aPD1_average_int_R <- colMeans(aPD1_average_int_R, na.rm = T)

colnames(aPD1_aIFNR_average_int_R) <- colnames(int_023_014_data)
aPD1_aIFNR_average_int_R <- colMeans(aPD1_aIFNR_average_int_R, na.rm = T)

colnames(hrIFNy_average_int_R) <- colnames(int_023_014_data)
hrIFNy_average_int_R <- colMeans(hrIFNy_average_int_R, na.rm = T)

#repeat for non-responders
unstim_average_int_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_int_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_int_NR <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_int_NR <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(int_list_NR)){
  data <- int_list_NR[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_int_NR <- rbind(unstim_average_int_NR, unstim)
    rownames(unstim_average_int_NR)[[nrow(unstim_average_int_NR)]] <- names(int_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_int_NR <- rbind(aPD1_average_int_NR, aPD1)
    rownames(aPD1_average_int_NR)[[nrow(aPD1_average_int_NR)]] <- names(int_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_int_NR <- rbind(aPD1_aIFNR_average_int_NR, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_int_NR)[[nrow(aPD1_aIFNR_average_int_NR)]] <- names(int_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_int_NR <- rbind(hrIFNy_average_int_NR, hrIFNy)
    rownames(hrIFNy_average_int_NR)[[nrow(hrIFNy_average_int_NR)]] <- names(int_list_NR)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_int_NR) <- colnames(int_023_014_data)
unstim_average_int_NR <- colMeans(unstim_average_int_NR, na.rm = T)

colnames(aPD1_average_int_NR) <- colnames(int_023_014_data)
aPD1_average_int_NR <- colMeans(aPD1_average_int_NR, na.rm = T)

colnames(aPD1_aIFNR_average_int_NR) <- colnames(int_023_014_data)
aPD1_aIFNR_average_int_NR <- colMeans(aPD1_aIFNR_average_int_NR, na.rm = T)

colnames(hrIFNy_average_int_NR) <- colnames(int_023_014_data)
hrIFNy_average_int_NR <- colMeans(hrIFNy_average_int_NR, na.rm = T)



#and tumour
tumour_list_R <- list(tumour_023_013_data,tumour_023_014_data,tumour_052_010_data, tumour_052_014_data,tumour_083_012_data, tumour_083_015_data)
names(tumour_list_R) <- c("LU023_TLS013","LU023_TLS014","LU052_TLS010","LU052_TLS014", "LU083_TLS012", "LU083_TLS015")  

tumour_list_NR <- list(tumour_035_013_data, tumour_035_015_data,tumour_066_012_data, tumour_066_015_data,tumour_080_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data)
names(tumour_list_NR) <- c("LU035_TLS013","LU035_TLS015","LU066_TLS012", "LU066_TLS015","LU080", "LU086-1", "LU086-2", "LU088", "LU091")  

unstim_average_tumour_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_tumour_R <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_tumour_R <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_tumour_R <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(tumour_list_R)){
  data <- tumour_list_R[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_tumour_R <- rbind(unstim_average_tumour_R, unstim)
    rownames(unstim_average_tumour_R)[[nrow(unstim_average_tumour_R)]] <- names(tumour_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_tumour_R <- rbind(aPD1_average_tumour_R, aPD1)
    rownames(aPD1_average_tumour_R)[[nrow(aPD1_average_tumour_R)]] <- names(tumour_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_tumour_R <- rbind(aPD1_aIFNR_average_tumour_R, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_tumour_R)[[nrow(aPD1_aIFNR_average_tumour_R)]] <- names(tumour_list_R)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_tumour_R <- rbind(hrIFNy_average_tumour_R, hrIFNy)
    rownames(hrIFNy_average_tumour_R)[[nrow(hrIFNy_average_tumour_R)]] <- names(tumour_list_R)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_tumour_R) <- colnames(tumour_023_014_data)
unstim_average_tumour_R <- colMeans(unstim_average_tumour_R, na.rm = T)

colnames(aPD1_average_tumour_R) <- colnames(tumour_023_014_data)
aPD1_average_tumour_R <- colMeans(aPD1_average_tumour_R, na.rm = T)

colnames(aPD1_aIFNR_average_tumour_R) <- colnames(tumour_023_014_data)
aPD1_aIFNR_average_tumour_R <- colMeans(aPD1_aIFNR_average_tumour_R, na.rm = T)

colnames(hrIFNy_average_tumour_R) <- colnames(tumour_023_014_data)
hrIFNy_average_tumour_R <- colMeans(hrIFNy_average_tumour_R, na.rm = T)

#repeat for non-responders
unstim_average_tumour_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_average_tumour_NR <- data.frame(matrix(nrow = 0, ncol = 25))
aPD1_aIFNR_average_tumour_NR <- data.frame(matrix(nrow = 0, ncol = 25))
hrIFNy_average_tumour_NR <- data.frame(matrix(nrow = 0, ncol = 25))

#Calculate condition averages per tumour
for(i in seq_along(tumour_list_NR)){
  data <- tumour_list_NR[[i]]
  
  unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
  aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
  aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
  hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
  
  unstim <- colMeans(unstim, na.rm = T)
  aPD1 <- colMeans(aPD1, na.rm = T)
  aPD1_aIFNR <- colMeans(aPD1_aIFNR, na.rm = T)
  hrIFNy <- colMeans(hrIFNy, na.rm = T)
  
  if(!any(grepl("NaN", unstim))){
    unstim_average_tumour_NR <- rbind(unstim_average_tumour_NR, unstim)
    rownames(unstim_average_tumour_NR)[[nrow(unstim_average_tumour_NR)]] <- names(tumour_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1))){
    aPD1_average_tumour_NR <- rbind(aPD1_average_tumour_NR, aPD1)
    rownames(aPD1_average_tumour_NR)[[nrow(aPD1_average_tumour_NR)]] <- names(tumour_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", aPD1_aIFNR))){
    aPD1_aIFNR_average_tumour_NR <- rbind(aPD1_aIFNR_average_tumour_NR, aPD1_aIFNR)
    rownames(aPD1_aIFNR_average_tumour_NR)[[nrow(aPD1_aIFNR_average_tumour_NR)]] <- names(tumour_list_NR)[[i]]
  }
  
  if(!any(grepl("NaN", hrIFNy))){
    hrIFNy_average_tumour_NR <- rbind(hrIFNy_average_tumour_NR, hrIFNy)
    rownames(hrIFNy_average_tumour_NR)[[nrow(hrIFNy_average_tumour_NR)]] <- names(tumour_list_NR)[[i]]
  }
  
}

#Average across all tumours
colnames(unstim_average_tumour_NR) <- colnames(tumour_023_014_data)
unstim_average_tumour_NR <- colMeans(unstim_average_tumour_NR, na.rm = T)

colnames(aPD1_average_tumour_NR) <- colnames(tumour_023_014_data)
aPD1_average_tumour_NR <- colMeans(aPD1_average_tumour_NR, na.rm = T)

colnames(aPD1_aIFNR_average_tumour_NR) <- colnames(tumour_023_014_data)
aPD1_aIFNR_average_tumour_NR <- colMeans(aPD1_aIFNR_average_tumour_NR, na.rm = T)

colnames(hrIFNy_average_tumour_NR) <- colnames(tumour_023_014_data)
hrIFNy_average_tumour_NR <- colMeans(hrIFNy_average_tumour_NR, na.rm = T)




#combine in a heatmap
#first: responder unstim-aPD1-combi
heatmap <- rbind(unstim_average_TLS_R,unstim_average_int_R,unstim_average_tumour_R,aPD1_average_TLS_R,aPD1_average_int_R,aPD1_average_tumour_R,
                 aPD1_aIFNR_average_TLS_R,aPD1_aIFNR_average_int_R,aPD1_aIFNR_average_tumour_R)

rownames(heatmap) <- c("unstim TLS" ,"unstim intermediate","unstim tumour bed","aPD1 TLS","aPD1 intermediate","aPD1 tumour bed",
                       "aPD1+aIFNyR1 TLS","aPD1+aIFNyR1 intermediate","aPD1+aIFNyR1 tumour bed")  

#z-scoring
colSD <- apply(heatmap, 2, sd)
colMean <- colMeans(heatmap)
z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y" #to make sure rownames can be seen
rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"

#heatmap
correlations <- pheatmap(z_scores, annotation_col = annotation,cluster_cols = F)

correlations

save_pheatmap(correlations, "~/Documents/Analyses/TLS/final_all_fragments/responder_triple.pdf")

#capture row order
row_order <- correlations$tree_row$order
values <- rownames(z_scores)[row_order]

#reorder z-scores based on row order
z_scores <- z_scores[order(match(rownames(z_scores), values)),]

#save z-scores
z_scores <- as.data.frame(z_scores)
write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/z_scores_responder_triple.xlsx", rowNames = T)

#second: responder unstim vs hrIFNy
heatmap <- rbind(unstim_average_TLS_R,unstim_average_int_R,unstim_average_tumour_R,hrIFNy_average_TLS_R,hrIFNy_average_int_R,hrIFNy_average_tumour_R)

rownames(heatmap) <- c("unstim TLS" ,"unstim intermediate","unstim tumour bed","hrIFNy TLS","hrIFNy intermediate","hrIFNy tumour bed")  

#z-scoring
colSD <- apply(heatmap, 2, sd)
colMean <- colMeans(heatmap)
z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y" #to make sure rownames can be seen
rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"

#heatmap
correlations <- pheatmap(z_scores, cluster_cols = F)

correlations

#capture row order
row_order <- correlations$tree_row$order
values <- rownames(z_scores)[row_order]

#reorder z-scores based on row order
z_scores <- z_scores[order(match(rownames(z_scores), values)),]

#save z-scores
z_scores <- as.data.frame(z_scores)
write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/z_scores_responder_double.xlsx", rowNames = T)

#third: non-responder unstim vs hrIFNy
heatmap <- rbind(unstim_average_TLS_NR,unstim_average_int_R,unstim_average_tumour_NR,hrIFNy_average_TLS_NR,hrIFNy_average_int_NR,hrIFNy_average_tumour_NR)

rownames(heatmap) <- c("unstim TLS" ,"unstim intermediate","unstim tumour bed","hrIFNy TLS","hrIFNy intermediate","hrIFNy tumour bed")  

#z-scoring
colSD <- apply(heatmap, 2, sd)
colMean <- colMeans(heatmap)
z_scores <- sweep(heatmap, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

rownames(z_scores)[rownames(z_scores) == "IFN-γ"] <- "IFN-y" #to make sure rownames can be seen
rownames(z_scores)[rownames(z_scores) == "TNF-α"] <- "TNF-a"

#heatmap
correlations <- pheatmap(z_scores, cluster_cols = F)

correlations

#capture row order
row_order <- correlations$tree_row$order
values <- rownames(z_scores)[row_order]

#reorder z-scores based on row order
z_scores <- z_scores[order(match(rownames(z_scores), values)),]

#save z-scores
z_scores <- as.data.frame(z_scores)
write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/z_scores_nonresponder_double.xlsx", rowNames = T)
```


Next, we calulate cumulative scores per condition and segment type for each fragment. The cumulative score is calculated as the sum of all z-scores per fragment. We can then compare segment types (TLS vs Int vs tumour in unstim and aPD1) as well as conditions (unstim vs aPD1 in TLS, int and tumour bed).
```{r Generate split data}
#Generate cumulative scores per fragment
#In this case I do not care about outliers as I want to compare the same amount of mediators for each fragment and removing one in a fragment lowers the count
  TLS_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% TLS_023_013,]
  int_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% int_023_013,]
  tumour_023_013_data <- lu023_TLS013[rownames(lu023_TLS013) %in% tumour_023_013,]
  
  TLS_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% TLS_023_014,]
  int_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% int_023_014,]
  tumour_023_014_data <- lu023_TLS014[rownames(lu023_TLS014) %in% tumour_023_014,]
  
  TLS_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% TLS_035_013,]
  int_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% int_035_013,]
  tumour_035_013_data <- lu035_TLS013[rownames(lu035_TLS013) %in% tumour_035_013,]
  
  TLS_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% TLS_035_015,]
  int_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% int_035_015,]
  tumour_035_015_data <- lu035_TLS015[rownames(lu035_TLS015) %in% tumour_035_015,]
  
  TLS_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% TLS_052_010,]
  int_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% int_052_010,]
  tumour_052_010_data <- lu052_TLS010[rownames(lu052_TLS010) %in% tumour_052_010,]
  #excluding IFNy_high fragments:
  TLS_052_010_data <- TLS_052_010_data[!grepl("hrIFNy_high", rownames(TLS_052_010_data)),]
  int_052_010_data <- int_052_010_data[!grepl("hrIFNy_high", rownames(int_052_010_data)),]
  tumour_052_010_data <- tumour_052_010_data[!grepl("hrIFNy_high", rownames(tumour_052_010_data)),]
  
  TLS_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% TLS_052_014,]
  int_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% int_052_014,]
  tumour_052_014_data <- lu052_TLS014[rownames(lu052_TLS014) %in% tumour_052_014,]
  
  #This tumour has no intermediate fragments. Therefore, this condition is skipped.
  TLS_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% TLS_066_012,]
  tumour_066_012_data <- lu066_TLS012[rownames(lu066_TLS012) %in% tumour_066_012,]
  
  TLS_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% TLS_066_015,]
  int_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% int_066_015,]
  tumour_066_015_data <- lu066_TLS015[rownames(lu066_TLS015) %in% tumour_066_015,]
  
  #LU080 has no TLS so we ignore this condition
  int_080_data <- lu080[rownames(lu080) %in% int_080,]
  tumour_080_data <- lu080[rownames(lu080) %in% tumour_080,]
  
  TLS_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% TLS_083_012,]
  int_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% int_083_012,]
  tumour_083_012_data <- lu083_TLS012[rownames(lu083_TLS012) %in% tumour_083_012,]
  
  TLS_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% TLS_083_015,]
  int_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% int_083_015,]
  tumour_083_015_data <- lu083_TLS015[rownames(lu083_TLS015) %in% tumour_083_015,]
  
  #the only TLS of LU086-1 were in RN440-15/RN440-16, which will not be considered. Therefore, we ignore this condition.
  int_086_1_data <- lu086_1[rownames(lu086_1) %in% int_086_1,]
  tumour_086_1_data <- lu086_1[rownames(lu086_1) %in% tumour_086_1,]
  
  #the only TLS of LU086-2 was in RN440-15, which was not taken along in the Legendplex. Therefore, we ignore this condition.
  int_086_2_data <- lu086_2[rownames(lu086_2) %in% int_086_2,]
  tumour_086_2_data <- lu086_2[rownames(lu086_2) %in% tumour_086_2,]
  
  TLS_088_data <- lu088[rownames(lu088) %in% TLS_088,]
  int_088_data <- lu088[rownames(lu088) %in% int_088,]
  tumour_088_data <- lu088[rownames(lu088) %in% tumour_088,]
  
  TLS_091_data <- lu091[rownames(lu091) %in% TLS_091,]
  int_091_data <- lu091[rownames(lu091) %in% int_091,]
  tumour_091_data <- lu091[rownames(lu091) %in% tumour_091,]
  
  
  #First for TLS
  TLS_list <- list(TLS_023_013_data,TLS_023_014_data, TLS_035_013_data, TLS_035_015_data, TLS_052_010_data, TLS_052_014_data, TLS_066_012_data,TLS_066_015_data,
                   TLS_083_012_data, TLS_083_015_data, TLS_088_data, TLS_091_data) #missing LU080, LU086-1, LU086-2
  names(TLS_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012","LU066_TLS015",
                       "LU083_TLS012", "LU083_TLS015", "LU088", "LU091")  
  
  unstim_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_TLS <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Separate per condition
  for(i in seq_along(TLS_list)){
    data <- TLS_list[[i]]
    
    
    if(grepl("TLS", names(TLS_list)[[i]])){ #if a tumour was spread over multiple experiments
      if(!grepl("LU035|LU052", names(TLS_list)[[i]])){ #and if that tumour was not present in the same legendplex (bc those already have the TLS number in front)
        number <- strsplit(names(TLS_list)[[i]], "_")
        number <- number[[1]][[2]]
        rownames(data) <- paste0(number, " ", rownames(data)) #add the experiment number in front of the rownames
      }
    }
    
    #split into separate conditions
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    #if the tumour has a certain condition, add it to the total
    if(nrow(unstim)!=0){
      unstim_average_TLS <- rbind(unstim_average_TLS, unstim)
      colnames(unstim_average_TLS) <- colnames(data)
    }
    
    if(nrow(aPD1)!=0){
      aPD1_average_TLS <- rbind(aPD1_average_TLS, aPD1)
      colnames(aPD1_average_TLS) <- colnames(data)
    }
    
    if(nrow(aPD1_aIFNR)!=0){
      aPD1_aIFNR_average_TLS <- rbind(aPD1_aIFNR_average_TLS, aPD1_aIFNR)
      colnames(aPD1_aIFNR_average_TLS) <- colnames(data)
    }
    
    if(nrow(hrIFNy)!=0){
      hrIFNy_average_TLS <- rbind(hrIFNy_average_TLS, hrIFNy)
      colnames(hrIFNy_average_TLS) <- colnames(data)
    }
    
  }
  

  #for intermediate
  int_list <- list(int_023_013_data,int_023_014_data, int_035_013_data, int_035_015_data, int_052_010_data, int_052_014_data,int_066_015_data, int_080_data,
                   int_083_012_data, int_083_015_data, int_086_1_data, int_086_2_data, int_088_data, int_091_data) #missing LU066
  names(int_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS015","LU080",
                       "LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
  
  unstim_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_int <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Separate per condition
  for(i in seq_along(int_list)){
    data <- int_list[[i]]
    
    
    if(grepl("TLS", names(int_list)[[i]])){ #if a tumour was spread over multiple experiments
      if(!grepl("LU035|LU052", names(int_list)[[i]])){ #and if that tumour was not present in the same legendplex (bc those already have the TLS number in front)
        number <- strsplit(names(int_list)[[i]], "_")
        number <- number[[1]][[2]]
        rownames(data) <- paste0(number, " ", rownames(data)) #add the experiment number in front of the rownames
      }
    }
    
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    
    if(nrow(unstim)!=0){
      unstim_average_int <- rbind(unstim_average_int, unstim)
      colnames(unstim_average_int) <- colnames(data)
    }
    
    if(nrow(aPD1)!=0){
      aPD1_average_int <- rbind(aPD1_average_int, aPD1)
      colnames(aPD1_average_int) <- colnames(data)
    }
    
    if(nrow(aPD1_aIFNR)!=0){
      aPD1_aIFNR_average_int <- rbind(aPD1_aIFNR_average_int, aPD1_aIFNR)
      colnames(aPD1_aIFNR_average_int) <- colnames(data)
    }
    
    if(nrow(hrIFNy)!=0){
      hrIFNy_average_int <- rbind(hrIFNy_average_int, hrIFNy)
      colnames(hrIFNy_average_int) <- colnames(data)
    }
    
  }
  
  #for tumour bed
  tumour_list <- list(tumour_023_013_data,tumour_023_014_data, tumour_035_013_data, tumour_035_015_data, tumour_052_010_data, tumour_052_014_data,tumour_066_012_data, tumour_066_015_data,
                      tumour_080_data,tumour_083_012_data, tumour_083_015_data, tumour_086_1_data, tumour_086_2_data, tumour_088_data, tumour_091_data) #missing LU066
  names(tumour_list) <- c("LU023_TLS013","LU023_TLS014","LU035_TLS013","LU035_TLS015","LU052_TLS010","LU052_TLS014","LU066_TLS012", "LU066_TLS015","LU080",
                       "LU083_TLS012", "LU083_TLS015","LU086-1", "LU086-2", "LU088", "LU091")  
  
  unstim_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  aPD1_aIFNR_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  hrIFNy_average_tumour <- data.frame(matrix(nrow = 0, ncol = 25))
  
  #Separate per condition. #NEED TO ADD EXPERIMENT NUMBER IN THERE SOMEHOW BC IT IS ADJUSTING FRAGMENT NAMING FOR ME WITH DUPLICATE ROWNAMES...
  for(i in seq_along(tumour_list)){
    data <- tumour_list[[i]]
    
    if(grepl("TLS", names(tumour_list)[[i]])){ #if a tumour was spread over multiple experiments
      if(!grepl("LU035|LU052", names(tumour_list)[[i]])){ #and if that tumour was not present in the same legendplex (bc those already have the TLS number in front)
        number <- strsplit(names(tumour_list)[[i]], "_")
        number <- number[[1]][[2]]
        rownames(data) <- paste0(number, " ", rownames(data)) #add the experiment number in front of the rownames
      }
    }
    
    unstim <- data[grepl("unstim", rownames(data), ignore.case = T),]
    aPD1 <- data[grepl("aPD1 fragment", rownames(data)),] #the "fragment" is there to prevent grabbing the combi condition
    aPD1_aIFNR <- data[grepl("aPD1\\+aIFNyR1", rownames(data)),]
    hrIFNy <- data[grepl("hrIFNy", rownames(data)),]
    
    
    if(nrow(unstim)!=0){
      unstim_average_tumour <- rbind(unstim_average_tumour, unstim)
      colnames(unstim_average_tumour) <- colnames(data)
    }
    
    if(nrow(aPD1)!=0){
      aPD1_average_tumour <- rbind(aPD1_average_tumour, aPD1)
      colnames(aPD1_average_tumour) <- colnames(data)
    }
    
    if(nrow(aPD1_aIFNR)!=0){
      aPD1_aIFNR_average_tumour <- rbind(aPD1_aIFNR_average_tumour, aPD1_aIFNR)
      colnames(aPD1_aIFNR_average_tumour) <- colnames(data)
    }
    
    if(nrow(hrIFNy)!=0){
      hrIFNy_average_tumour <- rbind(hrIFNy_average_tumour, hrIFNy)
      colnames(hrIFNy_average_tumour) <- colnames(data)
    }
    
  }
```

```{r Generate cumulative scores (unstim)}
#generate cumulative scores (unstim, all 3 segments)
#to separate between TLS from one tumour across two experiments, add the experiment number to the names
  #first, combine the data from all experiments and tumours
bind <- list(TLS_023_013,TLS_023_014, TLS_035_013, TLS_035_015, TLS_052_010, TLS_052_014, TLS_066_012, TLS_066_015, TLS_083_012, TLS_083_015, TLS_088, TLS_091,
             int_023_013,int_023_014, int_035_013, int_035_015, int_052_010, int_052_014, int_066_015, int_080, int_083_012, int_083_015, int_086_1, int_086_2, int_088, int_091,
             tumour_023_013,tumour_023_014, tumour_035_013, tumour_035_015, tumour_052_010, tumour_052_014, tumour_066_012, tumour_066_015, tumour_080, tumour_083_012, tumour_083_015, tumour_086_1, tumour_086_2, tumour_088, tumour_091)

#generate the names they should have
names(bind) <- c("TLS_023_013","TLS_023_014", "TLS_035_013", "TLS_035_015", "TLS_052_010", "TLS_052_014", "TLS_066_012", "TLS_066_015", "TLS_083_012", "TLS_083_015", "TLS_088", "TLS_091",
                 "int_023_013","int_023_014", "int_035_013", "int_035_015", "int_052_010", "int_052_014", "int_066_015", "int_080", "int_083_012", "int_083_015", "int_086_1", "int_086_2", "int_088", "int_091",
                 "tumour_023_013","tumour_023_014", "tumour_035_013", "tumour_035_015", "tumour_052_010", "tumour_052_014", "tumour_066_012", "tumour_066_015", "tumour_080", "tumour_083_012", "tumour_083_015", "tumour_086_1", "tumour_086_2", "tumour_088", "tumour_091")

#add the names to the fragments
for(i in seq_along(bind)){
    if(!grepl("035|052|080|086_1|086_2|088|091", names(bind)[[i]])){ #if a tumour was not present in the same legendplex (bc those already have the TLS number in front) or was not repeated, skip
      number <- strsplit(names(bind)[[i]], "_") #the TLS number is the second number in the name
      number <- number[[1]][[3]]
      bind[[i]] <- paste0("TLS", number, " ", bind[[i]]) #add the experiment number in front of the rownames
      
  }
}

#as we combined TLS, int and tumour for name-giving, split them up again
TLS_new <- unlist(bind[c(1:12)])
int_new <- unlist(bind[c(13:26)])
tumour_new <- unlist(bind[c(27:41)])

#calculate z-scores for the unstim condition
unstim_segment <- rbind(unstim_average_TLS,unstim_average_int,unstim_average_tumour)
colSD <- apply(unstim_segment, 2, sd, na.rm = T)
colMean <- colMeans(unstim_segment, na.rm = T)
z_scores <- sweep(unstim_segment, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"
cumulative_score$segment <- NA

#split fragments from the unstim condition into TLS, intermediate or tumour based on whether their name occurs in the TLS_new, int_new or tumour_new data frames
cumulative_score$segment[rownames(cumulative_score) %in% TLS_new] <- "TLS"
cumulative_score$segment[rownames(cumulative_score) %in% int_new] <- "Intermediate"
cumulative_score$segment[rownames(cumulative_score) %in% tumour_new] <- "Tumour bed"

#Add the tumour of origin
tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours #add to the data frame
cumulative_score$segment <- factor(cumulative_score$segment, levels = c("TLS", "Intermediate", "Tumour bed"))

#make a plot of cumulative scores in unstim; TLS vs int vs tumour
ggplot(cumulative_score, aes(x=segment, y= cumulative_score_fragment, col = tumour)) + geom_point() + ggtitle("unstim") + xlab("segment") + ylab("cumulative score") + theme_bw()

#export to prism
colnames(cumulative_score) <- c("value", "segment", "tumour")
cumulative_score$name <- rownames(cumulative_score)
test <- cumulative_score #since reshape does not like to have an additional column in there and cannot work with duplicated tumour names: remove it
test$tumour <- NULL
wide <- reshape(test, idvar = "name", timevar = "segment", direction = "wide") #convert to wide
wide$name <- NULL
colnames(wide) <- gsub("value\\.", "", colnames(wide)) #remove the "value" string that reshape adds to the column names
#since the row order is the same as the row order in the original data frame, we can just add the tumour name back in:
wide$tumours <- cumulative_score$tumour
#re-order columns
wide_unstim <- wide[,c(4,1,2,3)]

#export to excel
write.xlsx(wide_unstim, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_unstim_segments.xlsx",rowNames=T, colNames = T)
```

```{r Generate cumulative scores (aPD1)}
#repeat for aPD1
aPD1_segment <- rbind(aPD1_average_TLS,aPD1_average_int,aPD1_average_tumour)
colSD <- apply(aPD1_segment, 2, sd, na.rm = T)
colMean <- colMeans(aPD1_segment, na.rm = T)
z_scores <- sweep(aPD1_segment, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"
cumulative_score$segment <- NA

#split fragments from the aPD1 condition into TLS, intermediate or tumour based on whether their name occurs in the TLS_new, int_new or tumour_new data frames
cumulative_score$segment[rownames(cumulative_score) %in% TLS_new] <- "TLS" 
cumulative_score$segment[rownames(cumulative_score) %in% int_new] <- "Intermediate"
cumulative_score$segment[rownames(cumulative_score) %in% tumour_new] <- "Tumour bed"

#add tumour of origin
tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours
cumulative_score$segment <- factor(cumulative_score$segment, levels = c("TLS", "Intermediate", "Tumour bed"))
colnames(cumulative_score) <- c("value", "segment", "tumour")
cumulative_score$name <- rownames(cumulative_score)
test <- cumulative_score #since reshape does not like to have an additional column in there and cannot work with duplicated tumour names: remove it
test$tumour <- NULL
wide <- reshape(test, idvar = "name", timevar = "segment", direction = "wide") #convert to wide
wide$name <- NULL
colnames(wide) <- gsub("value\\.", "", colnames(wide))
#since the row order is the same as the row order in the original data frame, we can just add the tumour name back in:
wide$tumours <- cumulative_score$tumour
#re-order
wide_PD1 <- wide[,c(4,1,2,3)]

#export to excel
write.xlsx(wide_PD1, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_aPD1_segments.xlsx",rowNames=T, colNames = T)
```

```{r Calculate cumulative scores on a per-segment level}
#Calculate cumulative scores unstim vs aPD1 for each tumour
full <- rbind(unstim_segment, aPD1_segment) #combine all unstim and aPD1 fragments

#z-scoring
colSD <- apply(full, 2, sd, na.rm = T)
colMean <- colMeans(full, na.rm = T)
z_scores <- sweep(full, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)
cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"

#add the tumour of origin
tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours

#also add the condition of origin
cumulative_score$condition <- NA
cumulative_score$condition[grepl("unstim", rownames(cumulative_score))] <- "unstim"
cumulative_score$condition[grepl("PD1", rownames(cumulative_score))] <- "aPD1"
cumulative_score$tumour[grepl("LU086", rownames(cumulative_score))] <- "LU086" #fix the naming of LU086-1 and LU086-2; these are from the same tumour

#split the cumulative score per tumour
cumulative_score_tumour <- split(cumulative_score, cumulative_score$tumour)

#for each tumour, save the cumulative scores in unstim and aPD1 fragments to a csv file
for(i in seq_along(cumulative_score_tumour)){
  tumour <- names(cumulative_score_tumour)[[i]] #the tumour name is in the name of the list element
  cumulative_score_tumour[[i]]$tumour <- NULL
  cumulative_score_tumour[[i]]$name <- rownames(cumulative_score_tumour[[i]])
  
  wide <- reshape(cumulative_score_tumour[[i]], idvar = "name", timevar = "condition", direction = "wide") #convert to wide
  colnames(wide) <- c("fragment","unstim", "aPD1")
  
  write.csv(wide, paste0("~/Documents/Analyses/TLS/mass_data_export/", tumour, "_cumulative_score_unstim_aPD1.csv")) #export to csv
  
}
```

```{r Generate cumulative scores (TLS)}
#now compare unstim to aPD1 in the three different segments
#TLS
TLS_condition <- rbind(unstim_average_TLS,aPD1_average_TLS) #combine unstim and aPD1 TLS fragments

#z-scoring
colSD <- apply(TLS_condition, 2, sd, na.rm = T)
colMean <- colMeans(TLS_condition, na.rm = T)
z_scores <- sweep(TLS_condition, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"
cumulative_score$condition <- NULL

#assign fragments to unstim or aPD1 based on whether their name occurs in the unstim_TLS or aPD1_TLS data frames
cumulative_score$condition[rownames(cumulative_score) %in% rownames(unstim_average_TLS)] <- "unstim" #assign segments back to fragments based on whether they were in the segment-separated data frames
cumulative_score$condition[rownames(cumulative_score) %in% rownames(aPD1_average_TLS)] <- "aPD1"

#add te tumour of origin
tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours

cumulative_score$condition <- factor(cumulative_score$condition, levels = c("unstim", "aPD1"))
colnames(cumulative_score) <- c("value", "condition", "tumour")
cumulative_score$name <- rownames(cumulative_score)

#reshape into a wide format
test <- cumulative_score #since reshape does not like to have an additional column in there and cannot work with duplicated tumour names: remove it
test$tumour <- NULL
wide <- reshape(test, idvar = "name", timevar = "condition", direction = "wide") #convert to wide
wide$name <- NULL
colnames(wide) <- gsub("value\\.", "", colnames(wide))
#since the row order is the same as the row order in the original data frame, we can just add the tumour name back in:
wide$tumours <- cumulative_score$tumour
#re-order
wide_TLS <- wide[,c(3,1,2)]

#export to excel
write.xlsx(wide_TLS, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_TLS_conditions.xlsx",rowNames=T, colNames = T)
```

```{r Generate cumulative scores (Intermediate)}
#Repeat for intermediate fragments
int_condition <- rbind(unstim_average_int,aPD1_average_int)

colSD <- apply(int_condition, 2, sd, na.rm = T)
colMean <- colMeans(int_condition, na.rm = T)
z_scores <- sweep(int_condition, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"
cumulative_score$condition <- NULL

cumulative_score$condition[rownames(cumulative_score) %in% rownames(unstim_average_int)] <- "unstim" #assign segments back to fragments based on whether they were in the segment-separated data frames
cumulative_score$condition[rownames(cumulative_score) %in% rownames(aPD1_average_int)] <- "aPD1"
tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours

cumulative_score$condition <- factor(cumulative_score$condition, levels = c("unstim", "aPD1"))
colnames(cumulative_score) <- c("value", "condition", "tumour")
cumulative_score$name <- rownames(cumulative_score)

test <- cumulative_score #since reshape does not like to have an additional column in there and cannot work with duplicated tumour names: remove it
test$tumour <- NULL
wide <- reshape(test, idvar = "name", timevar = "condition", direction = "wide") #convert to wide
wide$name <- NULL
colnames(wide) <- gsub("value\\.", "", colnames(wide))
#since the row order is the same as the row order in the original data frame, we can just add the tumour name back in:
wide$tumours <- cumulative_score$tumour
#re-order
wide_int <- wide[,c(3,1,2)]

#export to excel
write.xlsx(wide_int, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_int_conditions.xlsx",rowNames=T, colNames = T)
```

```{r Generate cumulative scores (tumour bed)}
#And also for tumour bed
tumour_condition <- rbind(unstim_average_tumour,aPD1_average_tumour)

colSD <- apply(tumour_condition, 2, sd, na.rm = T)
colMean <- colMeans(tumour_condition, na.rm = T)
z_scores <- sweep(tumour_condition, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)

cumulative_score <- as.data.frame(colSums(z_scores)) #calculate cumulative scores by adding up z-scores per fragment
rownames(cumulative_score) <- colnames(z_scores)
colnames(cumulative_score) <- "cumulative_score_fragment"

cumulative_score$condition <- NULL
cumulative_score$condition[rownames(cumulative_score) %in% rownames(unstim_average_tumour)] <- "unstim" #assign segments back to fragments based on whether they were in the segment-separated data frames
cumulative_score$condition[rownames(cumulative_score) %in% rownames(aPD1_average_tumour)] <- "aPD1"

tumours <- str_extract_part(rownames(cumulative_score), before = FALSE, pattern = "LU") #grab the characters that come after "LU"
tumours <- strsplit(tumours, " ") #grab only the first characters that come after "LU" (which is the tumour number)
tumours <- sapply(tumours, "[[", 1) 
tumours <- paste0("LU", tumours) #put the "LU" back in front of the tumour number
cumulative_score$tumour <- tumours

cumulative_score$condition <- factor(cumulative_score$condition, levels = c("unstim", "aPD1"))
colnames(cumulative_score) <- c("value", "condition", "tumour")
cumulative_score$name <- rownames(cumulative_score)

test <- cumulative_score #since reshape does not like to have an additional column in there and cannot work with duplicated tumour names: remove it
test$tumour <- NULL
wide <- reshape(test, idvar = "name", timevar = "condition", direction = "wide") #convert to wide
wide$name <- NULL
colnames(wide) <- gsub("value\\.", "", colnames(wide))
#since the row order is the same as the row order in the original data frame, we can just add the tumour name back in:
wide$tumours <- cumulative_score$tumour
#re-order
wide_tumour <- wide[,c(3,1,2)]

#export to excel
write.xlsx(wide_tumour, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_tumour_conditions.xlsx",rowNames=T, colNames = T)
```

While we can generate cumulative scores for all tumours, it makes the most sense to look at differences between unstim and aPD1 in PDTF-R tumours. Therefore, we subset the cumulative scores generated in the previous sections to only include PDTF-R tumours.
```{r Repeat for PDTF-R only}
#repeat, but only take proven responders (LU023, LU052, LU083)
#From the calculation of the cumulative scores over all tumours, only grab the responder tumours (z-scores are not recalculated on the subsetted data frame)
wide_unstim_R <- wide_unstim[wide_unstim$tumours %in% c("LU023","LU052","LU083"),]
wide_PD1_R <- wide_PD1[wide_PD1$tumours %in% c("LU023","LU052","LU083"),]
wide_TLS_R <- wide_TLS[wide_TLS$tumours %in% c("LU023","LU052","LU083"),]
wide_int_R <- wide_int[wide_int$tumours %in% c("LU023","LU052","LU083"),]
wide_tumour_R <- wide_tumour[wide_tumour$tumours %in% c("LU023","LU052","LU083"),]

#export to excel
write.xlsx(wide_unstim_R, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_unstim_responder_segments.xlsx",rowNames=T, colNames = T)
write.xlsx(wide_PD1_R, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_aPD1_responder_segments.xlsx",rowNames=T, colNames = T)
write.xlsx(wide_TLS_R, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_TLS_responder_conditions.xlsx",rowNames=T, colNames = T)
write.xlsx(wide_int_R, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_int_responder_conditions.xlsx",rowNames=T, colNames = T)
write.xlsx(wide_tumour_R, "~/Documents/Analyses/TLS/final_all_fragments/cumulative_score_tumour_responder_conditions.xlsx",rowNames=T, colNames = T)
```

To see which mediators drive differences in cumulative scores, we can calculate the log2 fold-change between aPD1 and unstim (averaged across all tumours). First, we do this for all segment types combined and then we split it up into TLS, intermediate and tumour bed.
```{r Calculate log2FC aPD1/unstim}
#plot log2FC cytokine changes for responders
cytokines_unstim <- rbind(unstim_average_TLS, unstim_average_int, unstim_average_tumour) #grab unstim cytokine values
cytokines_aPD1 <- rbind(aPD1_average_TLS, aPD1_average_int, aPD1_average_tumour) #grab aPD1 cytokine values

cytokines_unstim <- cytokines_unstim[grepl("LU083|LU052|LU023", rownames(cytokines_unstim)),] #only take responders
cytokines_aPD1 <- cytokines_aPD1[grepl("LU083|LU052|LU023", rownames(cytokines_aPD1)),]

cytokines_unstim <- as.data.frame(colMeans(cytokines_unstim, na.rm = T)) #calculate averages per cytokine
cytokines_aPD1 <- as.data.frame(colMeans(cytokines_aPD1, na.rm = T)) #calculate averages per cytokine

colnames(cytokines_unstim) <- "unstim"
colnames(cytokines_aPD1) <- "aPD1"

#calculate the log2FC aPD1/unstim
cytokines <- cbind(cytokines_unstim, cytokines_aPD1)
cytokines$log2fc <- log2(cytokines_aPD1/cytokines_unstim)
colnames(cytokines) <- c("unstim","aPD1","log2fc")
cytokines[] <- lapply(cytokines, function(x) if(class(x) == 'data.frame') unlist(x) else x) #log2fc returns a column that is a data.frame somehow (class(cytokines[,3])) so change this
write.xlsx(cytokines, "~/Documents/Analyses/TLS/final_all_fragments/responder_average_cytokines_unstim_aPD1.xlsx",rowNames = T) #export to excel
```

```{r Calculate log2FC aPD1/unstim per segment type}
#separate by segment (they all have at least 2 fragments for each tumour)
unstim_average_TLS_small <- unstim_average_TLS[grepl("LU083|LU052|LU023", rownames(unstim_average_TLS)),] #only take responders
aPD1_average_TLS_small <- aPD1_average_TLS[grepl("LU083|LU052|LU023", rownames(aPD1_average_TLS)),] #only take responders

unstim_average_int_small <- unstim_average_int[grepl("LU083|LU052|LU023", rownames(unstim_average_int)),] #only take responders
aPD1_average_int_small <- aPD1_average_int[grepl("LU083|LU052|LU023", rownames(aPD1_average_int)),] #only take responders

unstim_average_tumour_small <- unstim_average_tumour[grepl("LU083|LU052|LU023", rownames(unstim_average_tumour)),] #only take responders
aPD1_average_tumour_small <- aPD1_average_tumour[grepl("LU083|LU052|LU023", rownames(aPD1_average_tumour)),] #only take responders

#calculate averages per condition and segment type for responders
unstim_average_TLS_small <- colMeans(unstim_average_TLS_small, na.rm = T) #calculate averages per cytokine
aPD1_average_TLS_small <- colMeans(aPD1_average_TLS_small, na.rm = T) #calculate averages per cytokine
unstim_average_int_small <- colMeans(unstim_average_int_small, na.rm = T) #calculate averages per cytokine
aPD1_average_int_small <- colMeans(aPD1_average_int_small, na.rm = T) #calculate averages per cytokine
unstim_average_tumour_small <- colMeans(unstim_average_tumour_small, na.rm = T) #calculate averages per cytokine
aPD1_average_tumour_small <- colMeans(aPD1_average_tumour_small, na.rm = T) #calculate averages per cytokine

#calculate the log2FC per segment type
TLS_log2FC <- log2(aPD1_average_TLS_small/unstim_average_TLS_small)
int_log2FC <- log2(aPD1_average_int_small/unstim_average_int_small)
tumour_log2FC <- log2(aPD1_average_tumour_small/unstim_average_tumour_small)

#combine and export
cytokines <- data.frame(cbind(TLS_log2FC, int_log2FC, tumour_log2FC))
colnames(cytokines) <- c("TLS","Intermediate","Tumour bed")
write.xlsx(cytokines, "~/Documents/Analyses/TLS/final_all_fragments/responder_log2FC_segment.xlsx",rowNames = T)
```

Next, we look at the heterogeneity between unstim and aPD1 fragments in our PDTF responders. We do this by generating a heatmap on a per-fragment level for each tumour.
```{r Individual PDTF-R heatmaps}
#show heterogeneity in LU083, LU052 and LU023 individual heatmaps (unstim and aPD1 only)
lu083 <- rbind(lu083_TLS012_copy, lu083_TLS015_copy) #use the copy versions as they also have experiment name in their fragment names: prevents duplicate fragment names and renaming
lu083 <- lu083[!grepl("average|hrIFNy|RN440-15|RN440-16|aPD1\\+aIFNyR1", rownames(lu083)),] #remove averages, RN440 conditions and the IFNy and aPD1+aIFNyR1 conditions

#z-scoring
colSD <- apply(lu083, 2, sd, na.rm = T)
colMean <- colMeans(lu083, na.rm = T)
z_scores <- sweep(lu083, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)
z_scores[z_scores == "NaN"] <- 0 #column SD is zero in some cases, likely due to normalisation

#rename vectors with fragment names for TLS etc. to match experiment tags
TLS_083_012_copy <- gsub("LU083", "LU083_TLS012", TLS_083_012)
TLS_083_015_copy <- gsub("LU083", "LU083_TLS015", TLS_083_015)

int_083_012_copy <- gsub("LU083", "LU083_TLS012", int_083_012)
int_083_015_copy <- gsub("LU083", "LU083_TLS015", int_083_015)

tumour_083_012_copy <- gsub("LU083", "LU083_TLS012", tumour_083_012)
tumour_083_015_copy <- gsub("LU083", "LU083_TLS015", tumour_083_015)

#reorder columns in this order: TLS-Intermediate-Tumour bed
TLS_083 <- z_scores[,colnames(z_scores) %in% c(TLS_083_012_copy, TLS_083_015_copy)]
int_083 <- z_scores[,colnames(z_scores) %in% c(int_083_012_copy, int_083_015_copy)]
tumour_083 <- z_scores[,colnames(z_scores) %in% c(tumour_083_012_copy, tumour_083_015_copy)]

#combine all fragments in the correct order (z-scores remain the same)
z_scores <- as.data.frame(cbind(TLS_083, int_083, tumour_083))

#make a heatmap
heatmap <- pheatmap(z_scores)

#reorder the z-scores based on the order of the heatmap (cytokine order)
row_order <- heatmap$tree_row$order
z_scores <- z_scores[order(row_order),]

#export the result
write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/lu083_z_scores.xlsx", rowNames= T)

#Repeat for LU052
lu052 <- rbind(lu052_TLS010, lu052_TLS014) #no need to add experiment names; they are already there
lu052 <- lu052[!grepl("average|hrIFNy|aPD1\\+aIFNyR1", rownames(lu052)),]
colSD <- apply(lu052, 2, sd, na.rm = T)
colMean <- colMeans(lu052, na.rm = T)
z_scores <- sweep(lu052, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)
z_scores[z_scores == "NaN"] <- 0 #column SD is zero in some cases, likely due to normalisation

#reorder columns in this order: TLS-Intermediate-Tumour bed
TLS_052 <- z_scores[,colnames(z_scores) %in% c(TLS_052_010, TLS_052_014)]
int_052 <- z_scores[,colnames(z_scores) %in% c(int_052_010, int_052_014)]
tumour_052 <- z_scores[,colnames(z_scores) %in% c(tumour_052_010, tumour_052_014)]

z_scores <- as.data.frame(cbind(TLS_052, int_052, tumour_052))
heatmap <- pheatmap(z_scores)
row_order <- heatmap$tree_row$order
z_scores <- z_scores[order(row_order),]

write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/lu052_z_scores.xlsx", rowNames= T)


#And also for LU023
lu023 <- rbind(lu023_TLS013_copy, lu023_TLS014_copy) #use the copy versions as they also have experiment name in their fragment names: prevents duplicate fragment names and renaming
lu023 <- lu023[!grepl("average|hrIFNy|RN440-15|RN440-16|aPD1\\+aIFNyR1", rownames(lu023)),]
colSD <- apply(lu023, 2, sd, na.rm = T)
colMean <- colMeans(lu023, na.rm = T)
z_scores <- sweep(lu023, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)
z_scores[z_scores == "NaN"] <- 0 #column SD is zero in some cases, likely due to normalisation

#rename vectors with fragment names for TLS etc. to match experiment tags
TLS_023_013_copy <- gsub("LU023", "LU023_TLS013", TLS_023_013)
TLS_023_014_copy <- gsub("LU023", "LU023_TLS014", TLS_023_014)

int_023_013_copy <- gsub("LU023", "LU023_TLS013", int_023_013)
int_023_014_copy <- gsub("LU023", "LU023_TLS014", int_023_014)

tumour_023_013_copy <- gsub("LU023", "LU023_TLS013", tumour_023_013)
tumour_023_014_copy <- gsub("LU023", "LU023_TLS014", tumour_023_014)

#reorder columns in this order: TLS-Intermediate-Tumour bed
TLS_023 <- z_scores[,colnames(z_scores) %in% c(TLS_023_013_copy, TLS_023_014_copy)]
int_023 <- z_scores[,colnames(z_scores) %in% c(int_023_013_copy, int_023_014_copy)]
tumour_023 <- z_scores[,colnames(z_scores) %in% c(tumour_023_013_copy, tumour_023_014_copy)]

z_scores <- as.data.frame(cbind(TLS_023, int_023, tumour_023))

heatmap <- pheatmap(z_scores)
row_order <- heatmap$tree_row$order
z_scores <- z_scores[order(row_order),]

write.xlsx(z_scores, "~/Documents/Analyses/TLS/final_all_fragments/lu023_z_scores.xlsx", rowNames= T)
```

```{r Session information}
sessionInfo()
```
