---
title: "Legendplex from template"
author: "Sidney van der Zande"
date: "`26-06-2024`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, collapse=T)
```

## Notes

This is a script meant to analyse Legendplex for the TLS experiments, using the accompanied Legendplex input template excel file. It only needs the tumour name, the path to the Legendplex file and an output directory to store the results in. The function outputs PDTF responder status for all fragments as well as all values it needs to calculate these scores such that they can be used later for other purposes.

The following section is to analyse one tumour at a time. If you want to analyse multiple tumours at a time, run the 'Main function' section and skip straight to the 'Batch input mode' section.

## Input

```{r Input data}
#Change the tumour name here
tumour_name <- "LU052"

#only use if the experiment number is important
#experiment <- NA
experiment = "TLS014"

#Give the FULL path to the Legendplex file
file_path <- "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/Final overview LP065.xlsx"

#In which directory should the output be stored?
output_dir <- "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/"

#For some functions, you will need this file
source("~/Documents/R scripts/Functions.R")
```

## Main function

If you have run this code block once, then you do not need to run it again and can just go to the 'Output' section directly (although running it again will not change anything)

```{r Main function: just run and do not change anything}
#function is defined here
analyze_legendplex <- function(tumour_name, file_path, batch_mode = F, experiment = NA){
  suppressMessages(require(readxl))
  
  #read in the file
  suppressMessages(file <- as.data.frame(read_excel(file_path, sheet = 2)))
  
  #failsafe for incorrect column names/numbers
  if(ncol(file) != 28){
    stop("Incorrect excel template")
  } 
  
  
  
  #set column names
  colnames(file) <- c("well", "sample", "IL-5", "IL-13", "IL-2", "IL-6", "IL-9", "IL-10", "IFN-γ", "TNF-α",
                      "IL-17A", "IL-17F", "IL-4", "IL-22", "empty", "IL-8", "CXCL10", "CCL11", "CCL17", "CCL2", "CCL5",
                      "CCL3", "CXCL9", "CXCL5", "CCL20", "CXCL1", "CXCL11", "CCL4")
  
  #Formatting:
  file$empty <- NULL
  file <- apply(file, 2, function(y) gsub("NA", NA, y)) #set excluded samples to NA
  file <- as.data.frame(file)
  file <- file[!is.na(file$sample),]
  file[,-c(1:2)] <- apply(file[,-c(1:2)], 2, function(y) as.numeric(y))
  
  
  #failsafe if PDTF cytokines are missing
  if(!(all(c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20") %in% colnames(file)))){
    missing_col <- colnames(file)[!(c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20") %in% colnames(file))]
    stop(paste0("Error: missing the following cytokines: ", missing_col))
  }
  
  #Grab the samples that correspond to the tumour name
  if(any(grepl(tumour_name, file$sample))){
    sample <- file[grepl(tumour_name, file$sample),]
  } else{
    stop("That tumour was not found in this file.")
  }
  
  if(!is.na(experiment)){
    sample <- sample[grepl(experiment, sample$sample),]
  }
  
  if(batch_mode == F){
    print(paste0("Results for tumour ", tumour_name))
    print(paste0(nrow(sample), " samples were found for that tumour."))
  }
  
  #Split into unstim and aPD1
  unstim <- sample[grepl("unstim", sample$sample, ignore.case = T),]
  rownames(unstim) <- unstim$sample
  unstim$sample <- NULL
  unstim$well <- NULL
  
  PD1 <- sample[grepl("aPD1", sample$sample, ignore.case = T),]
  PD1 <- PD1[!grepl("IFN", PD1$sample, ignore.case = F),] #exclude the aPD1+aIFNyR1 condition
  if(nrow(PD1) != 0){
    rownames(PD1) <- PD1$sample
    PD1 <- PD1[,!(colnames(PD1) %in% c("sample", "well"))]
    PD1_present <- T
  } else{
    PD1_present <- F
  }
  
  PD1_IFNR <- sample[grepl("aPD1\\+aIFNyR1", sample$sample, ignore.case = F),]
  if(nrow(PD1_IFNR) != 0){
    rownames(PD1_IFNR) <- PD1_IFNR$sample
    PD1_IFNR <- PD1_IFNR[,!(colnames(PD1_IFNR) %in% c("sample", "well"))]
    PD1_IFNR_present <- T
  } else{
    PD1_IFNR_present <- F
  }
  
  hrIFNy <- sample[grepl("hrIFNy", sample$sample, ignore.case = F),]
  if(nrow(hrIFNy) != 0){
    rownames(hrIFNy) <- hrIFNy$sample
    hrIFNy <- hrIFNy[,!(colnames(hrIFNy) %in% c("sample", "well"))]
    hrIFNy_present <- T
  } else{
    hrIFNy_present <- F
  }
  
  extra_conditions <- F
  
  if(grepl("LP060", file_path)){ #in LP060, samples were included where RN440-15 and RN440-16 conditions exist
    RN440_15 <- sample[grepl("RN440-15", sample$sample, ignore.case = F),]
    RN440_16 <- sample[grepl("RN440-16", sample$sample, ignore.case = F),]
    
    rownames(RN440_15) <- RN440_15$sample
    RN440_15 <- RN440_15[,!(colnames(RN440_15) %in% c("sample", "well"))]
    
    rownames(RN440_16) <- RN440_16$sample
    RN440_16 <- RN440_16[,!(colnames(RN440_16) %in% c("sample", "well"))]
    extra_conditions <- T
  }
  
  #get the blanco values
  blank <- file[grepl("Standard 0pg/ml", file$sample, ignore.case = T),]
  blank <- blank[,!(colnames(blank) %in% c("sample", "well"))]
  
  #average the values
  blank_avg <- colMeans(blank)
  
  #save uncorrected values
  uncorrected <- rbind(unstim, PD1, PD1_IFNR, hrIFNy)
  if(extra_conditions == T){
    uncorrected <- rbind(uncorrected, RN440_15, RN440_16)
  }
  uncorrected <- rbind(uncorrected, blank_avg)
  rownames(uncorrected)[length(rownames(uncorrected))] <- "Blank average" #change the name of the blank
  
  #correct for blancos in raw values
  unstim <- sweep(unstim, 2, blank_avg, "-")
  if(PD1_present == T){
    suppressWarnings(PD1 <- sweep(PD1, 2, blank_avg, "-"))
  }
  if(PD1_IFNR_present == T){
    suppressWarnings(PD1_IFNR <- sweep(PD1_IFNR, 2, blank_avg, "-"))
  }
  if(hrIFNy_present == T){
    suppressWarnings(hrIFNy <- sweep(hrIFNy, 2, blank_avg, "-"))
  }
  
  if(extra_conditions == T){
    suppressWarnings(RN440_15 <- sweep(RN440_15, 2, blank_avg, "-"))
    suppressWarnings(RN440_16 <- sweep(RN440_16, 2, blank_avg, "-"))
    
    RN440_15[RN440_15 < 0] <- 0
    RN440_15_avg <- colMeans(RN440_15)
    
    RN440_16[RN440_16 < 0] <- 0
    RN440_16_avg <- colMeans(RN440_16)
    
  }
  #set all negative values to zero
  unstim[unstim < 0] <- 0
  PD1[PD1 < 0] <- 0
  PD1_IFNR[PD1_IFNR < 0] <- 0
  hrIFNy[hrIFNy < 0] <- 0
  
  #calculate averages for stim and unstim
  unstim_avg <- colMeans(unstim)
  PD1_avg <- colMeans(PD1)
  PD1_IFNR_avg <- colMeans(PD1_IFNR)
  hrIFNy_avg <- colMeans(hrIFNy)
  
  #calculate the PD1 delta (PD1 - unstim average)
  if(PD1_present == T){
    suppressWarnings(PD1_delta <-  sweep(PD1, 2, unstim_avg, "-"))
  }
  
  #calculate unstim deltas (unstim - unstim average)
  unstim_delta <- sweep(unstim, 2, unstim_avg, "-")
  
  #calculate the aPD1+aIFNyR1 deltas (aPD1+aIFNyR1 - unstim average)
  if(PD1_IFNR_present == T){
    suppressWarnings(PD1_IFNR_delta <- sweep(PD1_IFNR, 2, unstim_avg, "-"))
  }
  
  #calculate the hrIFNy deltas (hrIFNy - unstim average)
  if(hrIFNy_present == T){
    suppressWarnings(hrIFNy_delta <- sweep(hrIFNy, 2, unstim_avg, "-"))
  }
  
  #calculate the RN440-15 and RN440-16 (RN440-15/16 - unstim average)
  if(extra_conditions == T){
    suppressWarnings(RN440_15_delta <- sweep(RN440_15, 2, unstim_avg, "-"))
    suppressWarnings(RN440_16_delta <- sweep(RN440_16, 2, unstim_avg, "-"))
  }
  
  #calculate PDTF scores (only if PD1 is present, though)
  if(PD1_present == T){
    PDTF_matrix_PD1 <- PD1_delta[,c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20")]
    PDTF_matrix_PD1[,1] <- ifelse(PDTF_matrix_PD1[,1] > 2.052, yes = 1, no = 0)
    PDTF_matrix_PD1[,2] <- ifelse(PDTF_matrix_PD1[,2] > 2.622, yes = 2, no = 0)
    PDTF_matrix_PD1[,3] <- ifelse(PDTF_matrix_PD1[,3] > 41.67, yes = 2, no = 0)
    PDTF_matrix_PD1[,4] <- ifelse(PDTF_matrix_PD1[,4] > 3.702, yes = 1, no = 0)
    PDTF_matrix_PD1[,5] <- ifelse(PDTF_matrix_PD1[,5] > 0.5597, yes = 2, no = 0)
    PDTF_matrix_PD1[,6] <- ifelse(PDTF_matrix_PD1[,6] > 395.3, yes = 1, no = 0)
    PDTF_matrix_PD1[,7] <- ifelse(PDTF_matrix_PD1[,7] > 482, yes = 2, no = 0)
    PDTF_matrix_PD1[,8] <- ifelse(PDTF_matrix_PD1[,8] > 0.3441, yes = 1, no = 0)
    PDTF_matrix_PD1[,9] <- ifelse(PDTF_matrix_PD1[,9] > 18.82, yes = 1, no = 0)
    PDTF_matrix_PD1[,10] <- ifelse(PDTF_matrix_PD1[,10] > 2.512, yes = 1, no = 0)
    PDTF_matrix_PD1[,11] <- ifelse(PDTF_matrix_PD1[,11] > 6.611, yes = 1, no = 0)
    
    #calculate the cumulative score, percentage and final responder status
    PDTF_matrix_PD1$total_score <- rowSums(PDTF_matrix_PD1)
    PDTF_matrix_PD1$percentage <- (PDTF_matrix_PD1$total_score/15)*100
    
    PDTF_matrix_PD1$PDTF_response <- ifelse(PDTF_matrix_PD1$percentage >= 40, yes = "PDTF-R", no = "PDTF-NR")
    PDTF_matrix_PD1$PDTF_response[PDTF_matrix_PD1$percentage >= 30 & PDTF_matrix_PD1$PDTF_response == "PDTF-NR"] <- "PDTF-BR"
    
    #calculate PDTF scores at baseline
    PDTF_matrix_unstim <- unstim_delta[,c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20")]
    PDTF_matrix_unstim[,1] <- ifelse(PDTF_matrix_unstim[,1] > 2.052, yes = 1, no = 0)
    PDTF_matrix_unstim[,2] <- ifelse(PDTF_matrix_unstim[,2] > 2.622, yes = 2, no = 0)
    PDTF_matrix_unstim[,3] <- ifelse(PDTF_matrix_unstim[,3] > 41.67, yes = 2, no = 0)
    PDTF_matrix_unstim[,4] <- ifelse(PDTF_matrix_unstim[,4] > 3.702, yes = 1, no = 0)
    PDTF_matrix_unstim[,5] <- ifelse(PDTF_matrix_unstim[,5] > 0.5597, yes = 2, no = 0)
    PDTF_matrix_unstim[,6] <- ifelse(PDTF_matrix_unstim[,6] > 395.3, yes = 1, no = 0)
    PDTF_matrix_unstim[,7] <- ifelse(PDTF_matrix_unstim[,7] > 482, yes = 2, no = 0)
    PDTF_matrix_unstim[,8] <- ifelse(PDTF_matrix_unstim[,8] > 0.3441, yes = 1, no = 0)
    PDTF_matrix_unstim[,9] <- ifelse(PDTF_matrix_unstim[,9] > 18.82, yes = 1, no = 0)
    PDTF_matrix_unstim[,10] <- ifelse(PDTF_matrix_unstim[,10] > 2.512, yes = 1, no = 0)
    PDTF_matrix_unstim[,11] <- ifelse(PDTF_matrix_unstim[,11] > 6.611, yes = 1, no = 0)
    
    #calculate the cumulative score, percentage and final responder status at baseline
    PDTF_matrix_unstim$total_score <- rowSums(PDTF_matrix_unstim)
    PDTF_matrix_unstim$percentage <- (PDTF_matrix_unstim$total_score/15)*100
    
    PDTF_matrix_unstim$PDTF_response <- ifelse(PDTF_matrix_unstim$percentage >= 40, yes = "PDTF-R", no = "PDTF-NR")
    PDTF_matrix_unstim$PDTF_response[PDTF_matrix_unstim$percentage >= 30 & PDTF_matrix_unstim$PDTF_response == "PDTF-NR"] <- "PDTF-BR"
    
    #calculate overall deltas
    overall_delta_PD1<- PD1_avg - unstim_avg
    
    #only keep cytokines we need
    responder_delta <- overall_delta_PD1[names(overall_delta_PD1) %in% c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20")]
    #correct order
    responder_delta <- responder_delta[order(factor(names(responder_delta), levels = c("IL-10", "IFN-γ", "CXCL10", "CCL5", "CCL17", "CXCL5", "CXCL1", "CXCL11", "CXCL9", "CCL4", "CCL20")))]
    
    #calculate overall responder scores
    responder_delta[1] <- ifelse(responder_delta[1] > 2.052, yes = 1, no = 0)
    responder_delta[2] <- ifelse(responder_delta[2] > 2.622, yes = 2, no = 0)
    responder_delta[3] <- ifelse(responder_delta[3] > 41.67, yes = 2, no = 0)
    responder_delta[4] <- ifelse(responder_delta[4] > 3.702, yes = 1, no = 0)
    responder_delta[5] <- ifelse(responder_delta[5] > 0.5597, yes = 2, no = 0)
    responder_delta[6] <- ifelse(responder_delta[6] > 395.3, yes = 1, no = 0)
    responder_delta[7] <- ifelse(responder_delta[7] > 482, yes = 2, no = 0)
    responder_delta[8] <- ifelse(responder_delta[8] > 0.3441, yes = 1, no = 0)
    responder_delta[9] <- ifelse(responder_delta[9] > 18.82, yes = 1, no = 0)
    responder_delta[10] <- ifelse(responder_delta[10] > 2.512, yes = 1, no = 0)
    responder_delta[11] <- ifelse(responder_delta[11] > 6.611, yes = 1, no = 0)
    
    #Sum the scores, calculate percentages and give overall responder status
    responder <- ((sum(responder_delta))/15)*100
    
    if(responder >= 40){
      responder_output <- "PDTF-R"
    } else if(responder >= 30){
      responder_output <- "PDTF-BR"
    } else{
      responder_output <- "PDTF-NR"
    }
    
    responder <- round(responder, 1)
    
  }
  
  #calculate the other overall deltas
  if(PD1_IFNR_present == T){
    overall_delta_PD1_IFNR<- PD1_IFNR_avg - unstim_avg
  }
  if(hrIFNy_present == T){
    overall_delta_hrIFNy<- hrIFNy_avg - unstim_avg
  }
  
  if(extra_conditions == T){
    overall_delta_RN440_15 <- RN440_15_avg - unstim_avg
    overall_delta_RN440_16 <- RN440_16_avg - unstim_avg
  }
  
  #return table with corrected cytokine values
  suppressWarnings(corrected_values <- rbind(unstim,PD1,PD1_IFNR,hrIFNy,unstim_avg,PD1_avg,PD1_IFNR_avg,hrIFNy_avg))
  if(PD1_present == T & PD1_IFNR_present ==F & hrIFNy_present == F & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1,unstim_avg,PD1_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "aPD1 average"
    
    deltas <- rbind(unstim_delta, PD1_delta, overall_delta_PD1)
    rownames(deltas)[length(rownames(deltas))] <- "overall delta aPD1"
    
  } else if(PD1_present == T & PD1_IFNR_present ==T & hrIFNy_present == F & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1,PD1_IFNR,unstim_avg,PD1_avg,PD1_IFNR_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-2] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "aPD1 average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "aPD1_aIFNyR1 average"
    
    deltas <- rbind(unstim_delta, PD1_delta, PD1_IFNR_delta, overall_delta_PD1,overall_delta_PD1_IFNR)
    rownames(deltas)[length(rownames(deltas))-1] <- "overall delta aPD1"
    rownames(deltas)[length(rownames(deltas))] <- "overall delta aPD1_aIFNR1"
    
  } else if(PD1_present == T & PD1_IFNR_present ==F & hrIFNy_present == T & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1,hrIFNy,unstim_avg,PD1_avg,hrIFNy_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-2] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "aPD1 average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "hrIFNy average"
    
    deltas <- rbind(unstim_delta, PD1_delta, hrIFNy_delta, overall_delta_PD1,overall_delta_hrIFNy)
    rownames(deltas)[length(rownames(deltas))-1] <- "overall delta aPD1"
    rownames(deltas)[length(rownames(deltas))] <- "overall delta hrIFNy"
    
  } else if(PD1_present == T & PD1_IFNR_present ==T & hrIFNy_present == T & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1,PD1_IFNR,hrIFNy,unstim_avg,PD1_avg,PD1_IFNR_avg,hrIFNy_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-3] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))-2] <- "aPD1 average"
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "aPD1_aIFNyR1 average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "hrIFNy average"
    
    deltas <- rbind(unstim_delta, PD1_delta, PD1_IFNR_delta, hrIFNy_delta, overall_delta_PD1,overall_delta_PD1_IFNR,overall_delta_hrIFNy)
    rownames(deltas)[length(rownames(deltas))-2] <- "overall delta aPD1"
    rownames(deltas)[length(rownames(deltas))-1] <- "overall delta aPD1_aIFNR1"
    rownames(deltas)[length(rownames(deltas))] <- "overall delta hrIFNy"
    
  } else if(PD1_present == F & PD1_IFNR_present ==F & hrIFNy_present == F & extra_conditions == F){
    corrected_values <- rbind(unstim,unstim_avg)
    rownames(corrected_values)[length(rownames(corrected_values))] <- "unstim average"
    
    deltas <- rbind(unstim_delta)
    
  } else if(PD1_present == F & PD1_IFNR_present ==T & hrIFNy_present == T & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1_IFNR,hrIFNy,unstim_avg,PD1_IFNR_avg,hrIFNy_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-2] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "aPD1_aIFNyR1 average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "hrIFNy average"
    
    deltas <- rbind(unstim_delta, PD1_IFNR_delta, hrIFNy_delta, overall_delta_PD1_IFNR,overall_delta_hrIFNy)
    rownames(deltas)[length(rownames(deltas))-1] <- "overall delta aPD1_aIFNR1"
    rownames(deltas)[length(rownames(deltas))] <- "overall delta hrIFNy"
    
  } else if(PD1_present == F & PD1_IFNR_present ==T & hrIFNy_present == F & extra_conditions == F){
    corrected_values <- rbind(unstim,PD1_IFNR,unstim_avg,PD1_IFNR_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "aPD1_aIFNyR1 average"
    
    deltas <- rbind(unstim_delta, PD1_IFNR_delta, overall_delta_PD1_IFNR)
    rownames(deltas)[length(rownames(deltas))] <- "overall delta aPD1_aIFNR1"
    
  } else if(PD1_present == F & PD1_IFNR_present ==F & hrIFNy_present == T & extra_conditions == F){
    corrected_values <- rbind(unstim,hrIFNy,unstim_avg,hrIFNy_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "hrIFNy average"
    
    deltas <- rbind(unstim_delta, hrIFNy_delta, overall_delta_hrIFNy)
    rownames(deltas)[length(rownames(deltas))] <- "overall delta hrIFNy"
  }
  
  #add RN440-15 and RN440-16 if applicable
  if(extra_conditions == T){ #extra conditions only exist if other 4 conditions exist too
    corrected_values <- rbind(unstim,PD1,PD1_IFNR,hrIFNy,RN440_15,RN440_16,unstim_avg,PD1_avg,PD1_IFNR_avg,hrIFNy_avg,RN440_15_avg,RN440_16_avg)
    rownames(corrected_values)[length(rownames(corrected_values))-5] <- "unstim average"
    rownames(corrected_values)[length(rownames(corrected_values))-4] <- "aPD1 average"
    rownames(corrected_values)[length(rownames(corrected_values))-3] <- "aPD1_aIFNyR1 average"
    rownames(corrected_values)[length(rownames(corrected_values))-2] <- "hrIFNy average"
    rownames(corrected_values)[length(rownames(corrected_values))-1] <- "RN440-15 average"
    rownames(corrected_values)[length(rownames(corrected_values))] <- "RN440-16 average"
    
    deltas <- rbind(unstim_delta, PD1_delta, PD1_IFNR_delta, hrIFNy_delta, RN440_15_delta, RN440_16_delta, overall_delta_PD1,overall_delta_PD1_IFNR,overall_delta_hrIFNy,overall_delta_RN440_15,overall_delta_RN440_16)
    rownames(deltas)[length(rownames(deltas))-4] <- "overall delta aPD1"
    rownames(deltas)[length(rownames(deltas))-3] <- "overall delta aPD1_aIFNR1"
    rownames(deltas)[length(rownames(deltas))-2] <- "overall delta hrIFNy"
    rownames(deltas)[length(rownames(deltas))-1] <- "overall delta RN440-15"
    rownames(deltas)[length(rownames(deltas))] <- "overall delta RN440-16"
    
  }
  
  
  #show responder status for individual fragments
  if(PD1_present == T){
    PDTF_matrix_full <- rbind(PDTF_matrix_unstim, PDTF_matrix_PD1)
    fragment_response <- data.frame(fragment = rownames(PDTF_matrix_full), response= PDTF_matrix_full$PDTF_response, percentage = PDTF_matrix_full$percentage)
    fragment_response$percentage <- round(fragment_response$percentage, 1)
    responder_total <- c(as.character(responder_delta), sum(responder_delta), as.character(responder), responder_output)
    PDTF_matrix_full <- rbind(PDTF_matrix_full, responder_total)
    rownames(PDTF_matrix_full)[length(rownames(PDTF_matrix_full))] <- "Overall"
    
    #produce output
    output <- list(uncorrected,corrected_values,deltas,PDTF_matrix_full)
    names(output) <- c("uncorrected values","corrected values","deltas","PDTF scores")
    
  } else{
    output <- list(uncorrected,corrected_values,deltas)
    names(output) <- c("uncorrected values","corrected values","deltas")
  }
  
  #Give output report:
  if(PD1_present == T){
    if(batch_mode == F){
      print("The fragments were classified as follows:")
      print(fragment_response)
    }
    
    print(paste0(tumour_name, " was overall classified as: ", responder_output, " (", responder, "%)"))
  }
  
  if(batch_mode == F){
    print("All values that were used for calculations can be found in the output of this function.")
  }
  
  return(output)
  
}
```

## Output

```{r Run the function, paged.print=FALSE}
#use the function, no changes are needed
output <- analyze_legendplex(file_path = file_path, tumour_name = tumour_name, experiment = experiment, batch_mode = F)
```
```{r Store the result}
#An excel file with the result (named the same as your tumour name) will be written to your output directory (or any other directory if you change the filename)
library(openxlsx)
write.xlsx(output, file=paste0(output_dir, tumour_name, "_", experiment, ".xlsx"), rowNames=T)
```

```{r Normalisation}
#Alternatively, you can process the file including normalisation of all values above or below the detection range.
#There are two ways to do this:
# normalisation = "windsor", which sets all values above the detectionlimit to he highest value below the detection limit and vice versa.
# normalisation = "percentile", which sets all values above the detection limit to the 99th percentile and vice versa.
source("~/Documents/R scripts/Functions.R")

LOD_file <- "~/Documents/Experiments/TLS/LP065-LU066-LU052(2)/LOD_LP065.xlsx"

output_normalised <- analyze_legendplex_v2(file_path = file_path, tumour_name = tumour_name, experiment = experiment, batch_mode = F, normalisation = "windsor", LOD_file = LOD_file)

library(openxlsx)
write.xlsx(output_normalised, file=paste0(output_dir, tumour_name, "_normalised_windsor.xlsx"), rowNames=T)

```



## Batch input mode

If you want to analyse multiple samples at a time (given that they appear in the same Legendplex file) then you can use this code chunk. It will not show PDTF status per fragment, only per tumour. The results are stored in the same way as the analysis per individual tumour above.

RUN THE 'Main function' SECTION BEFORE PROCEEDING

```{r Batch input}
#Change the tumour names here
tumour_name <- c("LU088", "LU091", "LU023") #if the same tumour from two experiments, input the tumour name once

#only fill this in if you use tumours from multiple experiments
#experiment <- c("TLS013", "TLS015")

#Give the FULL path to the Legendplex file, this file needs to contain all tumours entered above
file_path <- "~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/Final overview LP063.xlsx"

#In which directory should the output be stored?
output_dir <- "~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/"

#If performing normalisation, specify the LOD file here:
LOD_file <- "~/Documents/Experiments/TLS/LP063-LU088-LU091-LU023/LOD_LP063.xlsx"
```

```{r Run function}
#No changes needed, just run as is
library(openxlsx)

for(i in tumour_name){
  output <- analyze_legendplex(tumour_name = i, file_path = file_path, batch_mode = T)
  
write.xlsx(output, file=paste0(output_dir,i,".xlsx"), rowNames=T)
}
```


```{r Multiple experiments}
#Alternatively, if doing the same tumour from multiple experiments:
for(i in experiment){
  output <- analyze_legendplex(tumour_name = tumour_name, file_path = file_path, batch_mode = T, experiment = i)
  
write.xlsx(output, file=paste0(output_dir,i, tumour_name, ".xlsx"), rowNames=T)
}
```

```{r Batch mode normalisation}
#Alternatively, if normalisation needs to be applied:
source("~/Documents/R scripts/Functions.R")

for(i in tumour_name){
  output_normalised <- analyze_legendplex_v2(tumour_name = i, file_path = file_path, batch_mode = T, experiment = NA, normalisation = "windsor", LOD_file = LOD_file)
  
write.xlsx(output_normalised, file=paste0(output_dir,i,"_normalised_windsor.xlsx"), rowNames=T)
}
```

##Quality controls
Quality controls based on concentrations and MFIs have been moved to the Functions.R folder and used in the 'legendplex all tumours.R' file. Summon them from here:
```{r Quality control}
source("~/Documents/R scripts/Functions.R")
cat("Concentration:", "\n")
conc_qc <- legendplex_qc_concentration("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LU083.xlsx") #based on concentrations
conc_qc

cat("\n","MFI:","\n")
MFI_qc <- legendplex_qc_MFI("~/Documents/Experiments/TLS/LP060-RW_PK_01042024/LP060_MFI.xlsx", experiment = "TLS011", tumour_name = "LU083")
MFI_qc
```

##Plotting
```{r Make heatmaps}
#Generate a heatmap of z-scores per cytokine
library(pheatmap)

heatmap_list <- list()
experiment <- c("TLS014", "TLS016", "TLS016")

#prepare TLS status
TLS_status <- read.csv("~/Documents/Analyses/TLS/fragment_table.csv")
TLS_status$X <- NULL


for(i in tumour_name){
  
  index <- which(tumour_name == i)
  
  #read in TLS status
  TLS_status_i <- TLS_status[TLS_status$tumour == i & TLS_status$experiment == experiment[[index]],]
  TLS_status_i$heatmap <- paste0(TLS_status_i$tumour, " ", TLS_status_i$condition, " fragment " ,TLS_status_i$fragment)
  
  #read in the deltas
file_name <- paste0(output_dir,"/", i,".xlsx")
deltas <- read.xlsx(file_name, sheet = 3, colNames = T, rowNames = T) #sheet 3 contains delta values
deltas <- deltas[!grepl("overall",rownames(deltas)),]
values <- read.xlsx(file_name, sheet = 2, colNames = T, rowNames = T) #sheet 2 contains corrected values
average <- values[c("unstim average"),]
values <- values[!grepl("average", rownames(values)),]
  
  #z-score is (value - average)/standard deviation. (is this the delta or is the average per condition?)
colSD <- apply(values, 2, sd)
colMean <- colMeans(values)
z_scores <- sweep(values, 2, colMean, "-") #subtract means per column
z_scores <- sweep(z_scores, 2, colSD, "/") #divide by standard deviation per column
z_scores <- t(z_scores)


#exclude IFN-y as it might mess up the heatmap
#norm_mat <- norm_mat[rownames(norm_mat) != "IFN-γ",]

#add TLS status
colnames(z_scores) <- gsub("Unstim","unstim", colnames(z_scores))
TLS_status_i$heatmap <- gsub("aIFNyR1\\+aPD1", "aPD1+aIFNyR1", TLS_status_i$heatmap)
rownames(TLS_status_i) <- TLS_status_i$heatmap
TLS_status_i <- TLS_status_i[,c("heatmap","segment","cell_threshold_100")]
TLS_status_i <- TLS_status_i[rownames(TLS_status_i) %in% colnames(z_scores),] #missing LU023 fragment 6
TLS_status_i <- TLS_status_i[order(match(rownames(TLS_status_i), colnames(z_scores))),]
#not all fragments are in TLS data bc I could likely not find them during gating 
TLS_status_i$heatmap <- NULL

#set colours (TLS status)
cols <- list(segment = c(tumour_bed = "#99CCFF", intermediate = "#CC9FFF", TLS = "#99FFCC"), segment = c(YES = "pink", NO = "grey"))

#heatmap
heatmap <- suppressWarnings(pheatmap(z_scores, annotation_colors = cols, annotation_col = TLS_status_i))  

#save the heatmap
heatmap_list[[length(heatmap_list)+1]] <- heatmap
names(heatmap_list)[[length(heatmap_list)]] <- i

}

```
